### 2018-03-29 manuscript/main.tex
Initial commit of iSEE MS draft.

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

\begin{document}
\firstpage{1}

\subtitle{Data and text mining}

\title[iSEE]{iSEE: Interactive SummarizedExperiment Explorer}
\author[Rue-Albrecht \textit{et~al}.]{Kevin Rue-Albrecht\,$^{\text{\sfb 1}}$,
    Federico Marini\,$^{\text{\sfb 2}}$,
    Charlotte Soneson\,$^{\text{\sfb 3}}$
    and Aaron T. L. Lun\,$^{\text{\sfb 4,}*}$} 
\address{$^{\text{\sf 1}}$Kennedy Institute of Rheumatology, University of Oxford, Oxford OX3 7FY, United Kingdom, \\
$^{\text{\sf 2}}$Institute for Medical Biostatistics, Epidemiology and Informatics, \\
$^{\text{\sf 3}}$University of Zurich and SIB Swiss Institute of Bioinformatics, and \\
$^{\text{\sf 4}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Motivation:} Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
Text Text Text Text Text Text
Text Text Text Text Text.\\
\textbf{Results:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text\\
\textbf{Availability:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text Text  Text\\
\textbf{Contact:} \href{aaron.lun@cruk.cam.ac.uk}{aaron.lun@cruk.cam.ac.uk}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

\maketitle

\section{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics. 
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects, contaminants and low-quality samples. 
To this end, rapid visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, individual patients, single cells) and/or features (e.g., genes, transcripts, proteins, genomic regions).

Most existing tools for interactive visualization of biological data are designed for specific assays or analyses, e.g., pRoloc for proteomics data \citep{gatto2014mass}, shinyMethyl for methylation data \citep{fortin2014shinymethyl}, HTSvis for high-throughput screens \citep{scheeder2017htsvis}.
This makes it difficult to recycle the same visualization framework for new technologies or experimental designs where different aspects of the data are of interest. 
Moreover, standalone tools such as XXX do not easily integrate into established analysis frameworks such as those based on the R statistical programming language (CITE R Core Team). 
This complicates any coordinated use of these tools with a reproducible, transparent and rigorous statistical analysis.

\section{Description of the iSEE interface}
Here, we present the iSEE software package for interactive exploratory data analysis. 
iSEE is implemented in R using the interactive shiny framework and provides a number of features such as a customizable multi-panel layout, dynamic linking for point selection between plots, code reproducibility and intelligent downsampling. 
iSEE exploits the well-established SummarizedExperiment data architecture in the open-source Bioconductor project \citep{gentleman2004bioconductor}.
This enables iSEE to accommodate a wide range of `-omics datasets that are routinely processed by Bioconductor packages, ranging from conventional RNA sequencing to single-cell transcriptomics to mass cytometry.

\subsection{Integration with Bioconductor packages} 
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively. 
Per-feature or per-sample variables can also be easily added to the object in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results. 
The flexibility of the SummarizedExperiment class is the driving factor behind its wide deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, ChIP-seq \citep{lun2016csaw}, methylation \citep{aryee2014minfi}, Hi-C \citep{lun2016infrastructure} and mass cytometry (PMID: 28504682, lukas). 
Developers can easily derive subclasses for specialized use cases, such as the SingleCellExperiment class that enables easy storage of sample-wise reduced dimensionality embeddings in single-cell `-omics data analyses. 
By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization to complement existing analysis pipelines for a variety of data modalities. 
In particular, we leverage off the state-of-the-art statistical and computational methods already available in the R/Bioconductor framework, allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).

\subsection{Visualizing multiple data aspects simultaneously}
iSEE uses a customisable multi-panel layout that simultaneously displays one or more panels of different types, where each panel type visualizes a different aspect of the data. 
This includes metadata for each feature or sample, experimental observations across samples for each feature, and reduced dimensionality embeddings for each sample. 
For most panels, each sample or feature is represented as a point in a plot with the specified variables on the x/y-axes.
Depending on whether these variables are categorical or numeric, panels will automatically switch between scatter plots, violin plots or ``rectangle plots'' (where each combination of categorical levels is represented by a rectangle with area proportional to the number of samples/features with that combination). 
Each point can also be coloured according to data aspects such as sample-level metadata or feature-level observations. 
Other panel types include heatmaps, to visualize experimental observations for multiple features; and row statistics tables, to facilitate searching across features and their metadata.
All panels can be added, removed, reordered or resized.

\subsection{Dynamic linking across panels}
A key feature of iSEE is the ability to dynamically transmit information between panels. 
Users can define arbitrary links between a ``transmitting'' and ``receiving'' plot, whereby selection of points on the transmitting plot will highlight the corresponding points in the receiving plot (or optionally, hide all unselected points). 
This facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to a feature expression plot. 
The linking paradigm extends to multiple plots whereby a plot can transmit to multiple receivers, or a receiving plot can itself transmit to another plot.
This allows users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}, with the advantage of being extensible to any (meta)data present in a SummarizedExperiment object. 
Information can similarly be transferred between tables and plots, e.g., selecting a row determines how points are coloured in receiving plots.

\subsection{Customizing the user experience}
iSEE also offers a host of other customizable features to improve data exploration. 
Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each individual  metadata variable and assay type consistently across all panels in the application. 
Row statistics tables can be augmented with dynamic annotation based on the selected row, providing links to online resources such as Ensembl or Entrez annotation. 
For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization. 
After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.

\subsection{Reproducing the plotting code}
iSEE memorizes the exact R code that was used to generate every plot. 
This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts. 
All plots are generated using the ggplot2 framework and can be easily customized.

\section{Conclusion}
iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets. 
We provide a number of case studies at (REPOSITORY NAME) to demonstrate how iSEE can be used to explore the TCGA RNA-seq data set (CITE someone?), the 68,000 PBMC single-cell RNA sequencing dataset from 10X Genomics (CITE Zheng) and a published mass cytometry dataset consisting of over 170,000 cells (CITE?). 
Interested users can also easily apply iSEE on their own data by downloading the package from Bioconductor (https://bioconductor.org/packages/iSEE).

\section*{Acknowledgements}
We thank the organizers of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
We also thank members of the Bioconductor community for their helpful suggestions. 
\vspace*{-12pt}

\section*{Funding}
ATLL was supported by core funding from Cancer Research UK (award no. 17197 to Dr.\ John Marioni).
\vspace*{-12pt}

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
\bibliography{ref}

\end{document}

### 2018-03-29 manuscript/main.tex
Added abstract, tightened up some of the main text.

@@
 
-\abstract{\textbf{Motivation:} Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text Text
-Text Text Text Text Text Text
-Text Text Text Text Text.\\
-\textbf{Results:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text\\
-\textbf{Availability:} Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text Text  Text\\
+\abstract{\textbf{Summary:} Data exploration is criticial to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
+However, most existing tools for interactive visualization are limited to specific assays or analyses.
+Here, we present the iSEE (Interactive SummarizedExperiment Explorer) package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
+iSEE is directly compatible with many existing R/Bioconductor frameworks for analyzing high-throughput biological data,
+and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
+We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.\\
+\textbf{Availability:} iSEE is publicly available as an R package from the open-source Bioconductor project (https://bioconductor.org/packages/iSEE).\\
 \textbf{Contact:} \href{aaron.lun@cruk.cam.ac.uk}{aaron.lun@cruk.cam.ac.uk}\\
@@
 Per-feature or per-sample variables can also be easily added to the object in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results. 
-The flexibility of the SummarizedExperiment class is the driving factor behind its wide deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, ChIP-seq \citep{lun2016csaw}, methylation \citep{aryee2014minfi}, Hi-C \citep{lun2016infrastructure} and mass cytometry (PMID: 28504682, lukas).
-Developers can easily derive subclasses for specialized use cases, such as the SingleCellExperiment class that enables easy storage of sample-wise reduced dimensionality embeddings in single-cell `-omics data analyses.
+The flexibility of the SummarizedExperiment class is the driving factor behind its wide deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C \citep{lun2016infrastructure}.
+Developers can easily derive subclasses for specialized use cases, such as the SingleCellExperiment class that stores sample-wise reduced dimensionality embeddings in single-cell `-omics data analyses.
 By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization to complement existing analysis pipelines for a variety of data modalities. 
-In particular, we leverage off the state-of-the-art statistical and computational methods already available in the R/Bioconductor framework, allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).
+%In particular, we leverage off the state-of-the-art statistical and computational methods already available in the R/Bioconductor framework, allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).
 
 \subsection{Visualizing multiple data aspects simultaneously}
-iSEE uses a customisable multi-panel layout that simultaneously displays one or more panels of different types, where each panel type visualizes a different aspect of the data. 
+iSEE uses a customisable multi-panel layout that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data. 
 This includes metadata for each feature or sample, experimental observations across samples for each feature, and reduced dimensionality embeddings for each sample. 
@@
 Other panel types include heatmaps, to visualize experimental observations for multiple features; and row statistics tables, to facilitate searching across features and their metadata.
-All panels can be added, removed, reordered or resized.
+All panels can be removed, reordered or resized by the user, and new panels of any type can be added to the interface.
 
@@
 
-\subsection{Customizing the user experience}
+\subsection{Customizable and flexibility}
 iSEE also offers a host of other customizable features to improve data exploration. 
 Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each individual  metadata variable and assay type consistently across all panels in the application. 
-Row statistics tables can be augmented with dynamic annotation based on the selected row, providing links to online resources such as Ensembl or Entrez annotation. 
+Row statistics tables can be augmented with dynamic annotation based on the selected row, providing links to online resources such as Ensembl or Entrez. 
 For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
 After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
-
-\subsection{Reproducing the plotting code}
-iSEE memorizes the exact R code that was used to generate every plot. 
+iSEE also memorizes the exact R code that was used to generate every plot. 
 This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.
-All plots are generated using the ggplot2 framework and can be easily customized.
 

### 2018-03-29 manuscript/main.tex
Further nips and tucks, added relevant citations.

@@
 \copyrightyear{2015} \pubyear{2015}
-
 \access{Advance Access Publication Date: Day Month Year}
@@
 However, most existing tools for interactive visualization are limited to specific assays or analyses.
-Here, we present the iSEE (Interactive SummarizedExperiment Explorer) package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
-iSEE is directly compatible with many existing R/Bioconductor frameworks for analyzing high-throughput biological data,
+Here, we present the iSEE (Interactive SummarizedExperiment Explorer) software package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
+iSEE is directly compatible with many existing R/Bioconductor packages for analyzing high-throughput biological data,
 and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
@@
 Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics. 
-Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects, contaminants and low-quality samples.
-To this end, rapid visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, individual patients, single cells) and/or features (e.g., genes, transcripts, proteins, genomic regions).
+Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
+To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).
 
-Most existing tools for interactive visualization of biological data are designed for specific assays or analyses, e.g., pRoloc for proteomics data \citep{gatto2014mass}, shinyMethyl for methylation data \citep{fortin2014shinymethyl}, HTSvis for high-throughput screens \citep{scheeder2017htsvis}.
-This makes it difficult to recycle the same visualization framework for new technologies or experimental designs where different aspects of the data are of interest. 
-Moreover, standalone tools such as XXX do not easily integrate into established analysis frameworks such as those based on the R statistical programming language (CITE R Core Team). 
-This complicates any coordinated use of these tools with a reproducible, transparent and rigorous statistical analysis.
+Most existing tools for interactive visualization of biological data are designed for specific assays and analyses, e.g., pRoloc for proteomics \citep{gatto2014mass}, shinyMethyl for methylation \citep{fortin2014shinymethyl}, HTSvis for high-throughput screens \citep{scheeder2017htsvis}.
+Opportunities for customization are generally limited, making it difficult to re-use the same visualization software for new technologies or experimental designs where different aspects of the data are of interest.
+Moreover, standalone tools such as the Loupe Cell Browser from 10X Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}.
+This complicates any coordinated use of these tools with a reproducible, transparent and statistically rigorous analysis.
+
+Here, we present the iSEE software package for interactive data exploration.
+iSEE is implemented in R using the shiny framework \citep{chang2017shiny} and exploits the widely-used SummarizedExperiment data architecture from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
+This enables iSEE to accommodate a wide range of `-omics datasets that are routinely processed by Bioconductor packages, ranging from RNA sequencing to single-cell transcriptomics to mass cytometry.
 
 \section{Description of the iSEE interface}
-Here, we present the iSEE software package for interactive exploratory data analysis. 
-iSEE is implemented in R using the interactive shiny framework and provides a number of features such as a customizable multi-panel layout, dynamic linking for point selection between plots, code reproducibility and intelligent downsampling.
-iSEE exploits the well-established SummarizedExperiment data architecture in the open-source Bioconductor project \citep{gentleman2004bioconductor}.
-This enables iSEE to accommodate a wide range of `-omics datasets that are routinely processed by Bioconductor packages, ranging from conventional RNA sequencing to single-cell transcriptomics to mass cytometry.
 
@@
 This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively. 
-Per-feature or per-sample variables can also be easily added to the object in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results. 
-The flexibility of the SummarizedExperiment class is the driving factor behind its wide deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C \citep{lun2016infrastructure}.
-Developers can easily derive subclasses for specialized use cases, such as the SingleCellExperiment class that stores sample-wise reduced dimensionality embeddings in single-cell `-omics data analyses.
-By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization to complement existing analysis pipelines for a variety of data modalities. 
-%In particular, we leverage off the state-of-the-art statistical and computational methods already available in the R/Bioconductor framework, allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).
+Per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results. 
+The flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
+Developers can easily derive subclasses for specialized use cases, such as the SingleCellExperiment class that stores sample-wise reduced dimensionality embeddings for single-cell `-omics data analyses.
+By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in the R/Bioconductor framework.
+%In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).
 
@@
 
-\subsection{Customizable and flexibility}
+\subsection{Customizing the user experience}
 iSEE also offers a host of other customizable features to improve data exploration. 
@@
 iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets. 
-We provide a number of case studies at (REPOSITORY NAME) to demonstrate how iSEE can be used to explore the TCGA RNA-seq data set (CITE someone?), the 68,000 PBMC single-cell RNA sequencing dataset from 10X Genomics (CITE Zheng) and a published mass cytometry dataset consisting of over 170,000 cells (CITE?).
-Interested users can also easily apply iSEE on their own data by downloading the package from Bioconductor (https://bioconductor.org/packages/iSEE).
+We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection (CITE?), the 68,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells (CITE?).
 

### 2018-03-30 manuscript/pics/setup.R
Added script to setup iSEE for the screenshot.

@@ -1,0 +1,52 @@
+##############################
+# Setting up the Allen data. #
+##############################
+
+library(scRNAseq)
+data(allen)
+class(allen)
+
+library(scater)
+sce <- as(allen, "SingleCellExperiment")
+counts(sce) <- assay(sce, "tophat_counts")
+
+# Reduced dimensionality results.
+sce <- normalize(sce)
+sce <- runPCA(sce)
+
+# Adding feature-level metadata.
+rowData(sce)$ave_count <- rowMeans(counts(sce))
+rowData(sce)$n_cells <- rowSums(counts(sce)>0)
+
+library(org.Mm.eg.db)
+rowData(sce)$description <- mapIds(org.Mm.eg.db, keys=rownames(sce), keytype="SYMBOL", column="GENENAME")
+
+# # Choosing an interesting gene)
+# library(dynamicTreeCut)
+# mat <- dist(reducedDim(sce))
+# tree <- hclust(mat)
+# clusters <- cutreeDynamic(tree, distM=as.matrix(mat), minClusterSize=50)
+# library(scran)
+# markers <- findMarkers(sce, clusters)
+
+############################
+# Setting up iSEE options. #
+############################
+
+redDim <- DataFrame(ColorBy="Feature name", ColorByFeatName="Cux2",
+                    FontSize=1.5, PointSize=1.5)
+colData <- DataFrame(YAxis="NALIGNED", XAxis="Column data", XAxisColData="PRIMER",
+                     ColorBy="Column data", ColorByColData="passes_qc_checks_s",
+					 FontSize=1.5, PointSize=1.5)
+featExpr <- DataFrame(YAxisFeatName="Cux2", ColorBy="Column data", ColorByColData="driver_1_s",
+					 FontSize=1.5, PointSize=1.5)
+heatMap <- DataFrame(FeatName=I(list(c("Fam19a1", "Cacna2d3", "Cux2", "Rorb", "Ptn", "Lmo3", "Foxp2", "Lamp5", "Etv1"))),
+                     ColData=I(list(c("driver_1_s"))))
+rowStat <- DataFrame(Selected="Cux2", Search="homeobox 2")
+
+init <- data.frame(Name=c("Reduced dimension plot 1", "Column data plot 1", "Heat map 1", 
+                          "Row statistics table 1", "Feature expression plot 1"),
+                   Width=c(4,4,4,8,4))
+
+library(iSEE)
[…]

### 2018-03-30 manuscript/pics/setup.R
Updated Figure 1 to include screenshot, modified panel order.

@@ -44,8 +44,8 @@ heatMap <- DataFrame(FeatName=I(list(c("Fam19a1", "Cacna2d3", "Cux2", "Rorb", "P
                      ColData=I(list(c("driver_1_s"))))
 rowStat <- DataFrame(Selected="Cux2", Search="homeobox 2")
 
-init <- data.frame(Name=c("Reduced dimension plot 1", "Column data plot 1", "Heat map 1", 
-                          "Row statistics table 1", "Feature expression plot 1"),
+init <- data.frame(Name=c("Reduced dimension plot 1", "Column data plot 1", "Feature expression plot 1",
+                          "Row statistics table 1", "Heat map 1"),
                    Width=c(4,4,4,8,4))
 
 library(iSEE)

### 2018-03-30 manuscript/main.tex
Incorporated Figure 1, removed subheadings to save space.

@@
 
-Here, we present the iSEE software package for interactive data exploration.
-iSEE is implemented in R using the shiny framework \citep{chang2017shiny} and exploits the widely-used SummarizedExperiment data architecture from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
-This enables iSEE to accommodate a wide range of `-omics datasets that are routinely processed by Bioconductor packages, ranging from RNA sequencing to single-cell transcriptomics to mass cytometry.
+\begin{figure*}[t]
+\includegraphics[width=\textwidth]{pics/Fig1.pdf}
+\caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data. 
+New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
+Panel types are available visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v). 
+For these panel types, each sample is represented as a point, which can be coloured according to sample-level metadata or feature-level observations. 
+Depending on whether the variables of interest are categorical or numeric, panels will automatically switch between scatter plots, violin plots or ``rectangle plots'' (where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of combination).
+Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
+(B) Information can be transmitted between panels according to a user-specified scheme.
+Here, the selection of feature $X$ in the row statistics table determines the y-axis of the feature expression plot, and colours the samples in the reduced dimension plot by the expression of $X$.
+Selection of points in the reduced dimension plot (dotted blue line) also determines the samples that are shown in the column data (i.e., sample metadata) plot; 
+further selection of points in the column data plot determines the samples that are shown in the heatmap.
+}
+\label{fig:iSEE}
+\end{figure*}
 
 \section{Description of the iSEE interface}
-
-\subsection{Integration with Bioconductor packages} 
+Here, we present the iSEE software package for interactive data exploration.
+iSEE is implemented in R using the shiny framework \citep{chang2017shiny} and exploits data architectures from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
 The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
@@
 Per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results.
+All of these data aspects can be directly examined in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
 The flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
-Developers can easily derive subclasses for specialized use cases, such as the SingleCellExperiment class that stores sample-wise reduced dimensionality embeddings for single-cell `-omics data analyses.
-By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in the R/Bioconductor framework.
+By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in R/Bioconductor packages.
 %In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).
 
-\subsection{Visualizing multiple data aspects simultaneously}
-iSEE uses a customisable multi-panel layout that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data. 
-This includes metadata for each feature or sample, experimental observations across samples for each feature, and reduced dimensionality embeddings for each sample. 
-For most panels, each sample or feature is represented as a point in a plot with the specified variables on the x/y-axes.
-Depending on whether these variables are categorical or numeric, panels will automatically switch between scatter plots, violin plots or ``rectangle plots'' (where each combination of categorical levels is represented by a rectangle with area proportional to the number of samples/features with that combination).
-Each point can also be coloured according to data aspects such as sample-level metadata or feature-level observations. 
-Other panel types include heatmaps, to visualize experimental observations for multiple features; and row statistics tables, to facilitate searching across features and their metadata.
-All panels can be removed, reordered or resized by the user, and new panels of any type can be added to the interface.
-
-\subsection{Dynamic linking across panels}
-A key feature of iSEE is the ability to dynamically transmit information between panels. 
-Users can define arbitrary links between a ``transmitting'' and ``receiving'' plot, whereby selection of points on the transmitting plot will highlight the corresponding points in the receiving plot (or optionally, hide all unselected points).
+A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
+Users can define arbitrary links between ``transmitting'' and ``receiving'' plots, whereby selection of points on the transmitting plot will highlight the corresponding points in the receiving plot. 
 This facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to a feature expression plot.
@@
 
-\subsection{Customizing the user experience}
 iSEE also offers a host of other customizable features to improve data exploration. 
-Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each individual  metadata variable and assay type consistently across all panels in the application. 
-Row statistics tables can be augmented with dynamic annotation based on the selected row, providing links to online resources such as Ensembl or Entrez. 
+Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each metadata variable and assay type across all panels in the application.
+Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez. 
 For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
@@
 \section*{Acknowledgements}
-We thank the organizers of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
+We thank the organizers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
 We also thank members of the Bioconductor community for their helpful suggestions. 

### 2018-04-03 manuscript/main.tex
added my other affiliation, plus funding info

@@
 \author[Rue-Albrecht \textit{et~al}.]{Kevin Rue-Albrecht\,$^{\text{\sfb 1}}$,
-    Federico Marini\,$^{\text{\sfb 2}}$,
-    Charlotte Soneson\,$^{\text{\sfb 3}}$
-    and Aaron T. L. Lun\,$^{\text{\sfb 4,}*}$} 
+    Federico Marini\,$^{\text{\sfb 2}}$,$^{\text{\sfb 3}}$,
+    Charlotte Soneson\,$^{\text{\sfb 4}}$
+    and Aaron T. L. Lun\,$^{\text{\sfb 5,}*}$} 
 \address{$^{\text{\sf 1}}$Kennedy Institute of Rheumatology, University of Oxford, Oxford OX3 7FY, United Kingdom, \\
-$^{\text{\sf 2}}$Institute for Medical Biostatistics, Epidemiology and Informatics, \\
-$^{\text{\sf 3}}$University of Zurich and SIB Swiss Institute of Bioinformatics, and \\
-$^{\text{\sf 4}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom}
+$^{\text{\sf 2}}$Center for Thrombosis and Hemostasis (CTH), Mainz, \\
+$^{\text{\sf 3}}$Institute for Medical Biostatistics, Epidemiology and Informatics, Mainz, \\
+$^{\text{\sf 4}}$University of Zurich and SIB Swiss Institute of Bioinformatics, and \\
+$^{\text{\sf 5}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom}
 
@@
 ATLL was supported by core funding from Cancer Research UK (award no. 17197 to Dr.\ John Marioni).
+The work of FM is supported by the German Federal Ministry of Education and Research (BMBF 01EO1003).
 \vspace*{-12pt}

### 2018-04-03 manuscript/main.tex
small typo, plus couple of comments

@@
 
-\abstract{\textbf{Summary:} Data exploration is criticial to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
+\abstract{\textbf{Summary:} Data exploration is critical to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
 However, most existing tools for interactive visualization are limited to specific assays or analyses.
@@
 Moreover, standalone tools such as the Loupe Cell Browser from 10X Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}.
-This complicates any coordinated use of these tools with a reproducible, transparent and statistically rigorous analysis.
+This complicates any coordinated use of these tools with a reproducible, transparent, and statistically rigorous analysis.
 
@@
 Users can define arbitrary links between ``transmitting'' and ``receiving'' plots, whereby selection of points on the transmitting plot will highlight the corresponding points in the receiving plot. 
-This facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to a feature expression plot.
+This facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to one or more feature expression plots.
 The linking paradigm extends to multiple plots whereby a plot can transmit to multiple receivers, or a receiving plot can itself transmit to another plot.
@@
 Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez.
+% do we need to cite them?
+at least ENSEMBL has updated citable papers
 For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.

### 2018-04-03 manuscript/main.tex
slight edits to sentence for repro, plus adding extra citep

@@
 After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
-iSEE also memorizes the exact R code that was used to generate every plot. 
+iSEE also automatically memorizes the exact R code that was used to generate every plot and the existing links, 
+enhancing the proposal of \citep{marini2016interrepro}, where a template report was able to capture the live status of the web application after user interaction.
 This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.

### 2018-04-04 manuscript/main.tex
compacted affiliation environment commands

@@
 \author[Rue-Albrecht \textit{et~al}.]{Kevin Rue-Albrecht\,$^{\text{\sfb 1}}$,
-    Federico Marini\,$^{\text{\sfb 2}}$,$^{\text{\sfb 3}}$,
-    Charlotte Soneson\,$^{\text{\sfb 4}}$
+    Federico Marini\,$^{\text{\sfb 2,3}}$,
+    Charlotte Soneson\,$^{\text{\sfb 4}}$,
     and Aaron T. L. Lun\,$^{\text{\sfb 5,}*}$} 

### 2018-04-04 manuscript/main.tex
cut to the barebone size the extra ref

@@
 iSEE also automatically memorizes the exact R code that was used to generate every plot and the existing links, 
-enhancing the proposal of \citep{marini2016interrepro}, where a template report was able to capture the live status of the web application after user interaction.
+extending previous work by \cite{marini2016interrepro}.
 This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.

### 2018-04-07 manuscript/main.tex

@@
 iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
-We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection (CITE?), the 68,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells (CITE?).
+We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 68,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells (CITE?).
 

### 2018-04-10 tours/cytof_tour_code.R
Add CyTOF tour

@@ -1,0 +1,54 @@
+## Read data object and tour steps
+tour_steps <- read.delim("isee_cytof_tour.txt" header = TRUE, as.is = TRUE, sep = ";")
+load(url("https://www.dropbox.com/s/ogqp5grxe39c02d/bcrxl_t.rda?dl=1"))  ## bcrxl object
+
+## Packages 
+stopifnot(require(iSEE))
+
+## Set up panels (6 feature expression plots)
+initialPanels = DataFrame(
+    Name = c(sprintf("Feature expression plot %i", seq(1, nrow(fd), 1))),
+    Width = 4)
+
+fd <- featExprPlotDefaults(bcrxl, 6)
+
+## Appearance
+fd$VisualBoxOpen <- c(TRUE, FALSE, FALSE, FALSE, FALSE, TRUE)
+fd$DataBoxOpen <- c(TRUE, FALSE, FALSE, FALSE, FALSE, FALSE)
+fd$SelectBoxOpen <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE)
+fd$VisualChoices <- lapply(seq_len(nrow(fd)), function(i) list("Color", "Points"))
+
+## X-axis
+fd$XAxis <- rep(c("Column data", "Feature name", "Column data"), c(3, 2, 1))
+fd$XAxisColData <- rep(c("sample_id", "group_id"), c(5, 1))
+fd$XAxisFeatName <- match(c("", "", "", "CD3", "IgM", ""), rownames(bcrxl))
+
+## Y-axis
+fd$YAxisFeatName <- match(c("BC2", "BC4", "BC6", "CD20", "CD45", "pS6"), rownames(bcrxl))
+
+## Coloring
+fd$ColorBy <- c(rep("Column data", 3), "None", "None", "Column data")
+fd$ColorByColData <- c(rep("sample_id", 3), rep("group_id", 3))
+
+## Downsampling
+fd$Downsample <- TRUE
+fd$SampleRes <- 100
+
+## Point selection
+fd$SelectByPlot <- c("", "", "", "", "Feature expression plot 4", "Feature expression plot 5")
+fd$SelectEffect <- "Restrict"
+fd$BrushData <- list(NULL, NULL, NULL, 
[…]

### 2018-04-10 tours/cytof_tour_code.R
Add app title

@@ -49,6 +49,7 @@ app <- iSEE(bcrxl, tour = tour_steps,
             redDimMax = 0, colDataMax = 0,
             featExprMax = nrow(fd), 
             rowStatMax = 0, rowDataMax = 0, heatMapMax = 0,
-            initialPanels = initialPanels)
+            initialPanels = initialPanels,
+            appTitle = "CyTOF data tour")
 
 shiny::runApp(app)

### 2018-04-12 manuscript/main.tex
authors contributed equally

@@
 \title[iSEE]{iSEE: Interactive SummarizedExperiment Explorer}
-\author[Rue-Albrecht \textit{et~al}.]{Kevin Rue-Albrecht\,$^{\text{\sfb 1}}$,
-    Federico Marini\,$^{\text{\sfb 2,3}}$,
-    Charlotte Soneson\,$^{\text{\sfb 4}}$,
-    and Aaron T. L. Lun\,$^{\text{\sfb 5,}*}$}
+\author[Rue-Albrecht \textit{et~al}.]{Kevin Rue-Albrecht\,$^{\text{\sfb 1,}\dagger}$,
+    Federico Marini\,$^{\text{\sfb 2,3,}\dagger}$,
+    Charlotte Soneson\,$^{\text{\sfb 4,}\dagger}$,
+    and Aaron T. L. Lun\,$^{\text{\sfb 5,}\ast,\dagger}$}
 \address{$^{\text{\sf 1}}$Kennedy Institute of Rheumatology, University of Oxford, Oxford OX3 7FY, United Kingdom, \\
-$^{\text{\sf 2}}$Center for Thrombosis and Hemostasis (CTH), Mainz, \\
-$^{\text{\sf 3}}$Institute for Medical Biostatistics, Epidemiology and Informatics, Mainz, \\
-$^{\text{\sf 4}}$University of Zurich and SIB Swiss Institute of Bioinformatics, and \\
-$^{\text{\sf 5}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom}
+$^{\text{\sf 2}}$Center for Thrombosis and Hemostasis (CTH), Mainz \\
+$^{\text{\sf 3}}$Institute for Medical Biostatistics, Epidemiology and Informatics, Mainz \\
+$^{\text{\sf 4}}$University of Zurich and SIB Swiss Institute of Bioinformatics \\
+$^{\text{\sf 5}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom \\
+$^{^\dagger}$These authors contributed equally.}
 

### 2018-04-12 manuscript/main.tex
minor tweak

@@
 \address{$^{\text{\sf 1}}$Kennedy Institute of Rheumatology, University of Oxford, Oxford OX3 7FY, United Kingdom, \\
-$^{\text{\sf 2}}$Center for Thrombosis and Hemostasis (CTH), Mainz \\
-$^{\text{\sf 3}}$Institute for Medical Biostatistics, Epidemiology and Informatics, Mainz \\
-$^{\text{\sf 4}}$University of Zurich and SIB Swiss Institute of Bioinformatics \\
-$^{\text{\sf 5}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom \\
+$^{\text{\sf 2}}$Center for Thrombosis and Hemostasis (CTH), Mainz, \\
+$^{\text{\sf 3}}$Institute for Medical Biostatistics, Epidemiology and Informatics, Mainz, \\
+$^{\text{\sf 4}}$University of Zurich and SIB Swiss Institute of Bioinformatics, \\
+$^{\text{\sf 5}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom, and \\
 $^{^\dagger}$These authors contributed equally.}

### 2018-04-12 manuscript/main.tex
last typo (I hope)

--- 2018-04-12 manuscript/main.tex
@@
 \address{$^{\text{\sf 1}}$Kennedy Institute of Rheumatology, University of Oxford, Oxford OX3 7FY, United Kingdom, \\
-$^{\text{\sf 2}}$Center for Thrombosis and Hemostasis (CTH), Mainz \\
-$^{\text{\sf 3}}$Institute for Medical Biostatistics, Epidemiology and Informatics, Mainz \\
-$^{\text{\sf 4}}$University of Zurich and SIB Swiss Institute of Bioinformatics \\
-$^{\text{\sf 5}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom \\
+$^{\text{\sf 2}}$Center for Thrombosis and Hemostasis (CTH), Mainz, \\
+$^{\text{\sf 3}}$Institute for Medical Biostatistics, Epidemiology and Informatics, Mainz, \\
+$^{\text{\sf 4}}$University of Zurich and SIB Swiss Institute of Bioinformatics, and \\
+$^{\text{\sf 5}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom\\
 $^{^\dagger}$These authors contributed equally.}

### 2018-04-12 manuscript/main.tex
generalisation of plot -> panel; and data points

@@
 A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
-Users can define arbitrary links between ``transmitting'' and ``receiving'' plots, whereby selection of points on the transmitting plot will highlight the corresponding points in the receiving plot.
+Users can define arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting tables and plots highlight the corresponding objects in receiving panels.
 This facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to one or more feature expression plots.

### 2018-04-12 manuscript/main.tex
dropped a word to sneak the funding entirely in the second page. only references in 3rd page

@@
 A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
-Users can define arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting tables and plots highlight the corresponding objects in receiving panels.
+Users can define arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting tables and plots highlight the corresponding data points in receiving panels.
 This facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to one or more feature expression plots.
@@
 \section*{Acknowledgements}
-We thank the organizers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
-We also thank members of the Bioconductor community for their helpful suggestions.
+We thank organizers and participants at the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
+We also thank members of the Bioconductor community for helpful suggestions.
 \vspace*{-12pt}

### 2018-04-13 manuscript/main.tex
changed back

@@
 \section*{Acknowledgements}
-We thank organizers and participants at the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
-We also thank members of the Bioconductor community for helpful suggestions.
+We thank the organizers and participants at the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
+We also thank members of the Bioconductor community for their helpful suggestions.
 \vspace*{-12pt}

### 2018-04-16 manuscript/main.tex

@@
 \section*{Acknowledgements}
-We thank the organizers and participants at the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
+We thank the organizers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
 We also thank members of the Bioconductor community for their helpful suggestions.

### 2018-04-16 tours/pbmc4k_app.R
Added 4K PBMC tour based on emptyDrops paper.

@@ -1,0 +1,45 @@
+library(iSEE)
+library(SingleCellExperiment)
+
+# Obtain an SCE object by running the vignette at:
+# https://github.com/MarioniLab/EmptyDrops2017/tree/master/analysis/pbmc4k/analysis.Rmd
+sce <- readRDS("sce.rds")
+
+# Setting up defaults for all panels.
+redDim <- DataFrame(Type="TSNE", ColorBy="Column data", ColorByColData="Detection")
+colData <- DataFrame(YAxis="log10_total_counts", XAxis="Column data", XAxisColData="Cluster", ColorBy="Column data", ColorByColData="Detection")
+featExpr <- DataFrame(YAxisFeatName="PF4", XAxis="Column data", XAxisColData="Cluster", ColorBy="Column data", ColorByColData="log10_total_counts")
+heatMap <- DataFrame(FeatName=I(list(c("PF4", "PPBP", "TMSB4X", "B2M", "ACTB", "SDPR"))), 
+                     ColData=I(list(c("Cluster"))), Lower=-5, Upper=5)
+rowStat <- DataFrame(Search="PF4", Selected="PF4")
+
+# Setting up links between plots.
+colData$SelectByPlot <- "Reduced dimension plot 1"
+colData$SelectEffect <- "Transparent"
+colData$SelectAlpha <- 0
+
+featExpr$SelectByPlot <- "Reduced dimension plot 1"
+featExpr$SelectEffect <- "Transparent"
+featExpr$SelectAlpha <- 0
+featExpr$YAxisRowTable <- "Row statistics table 1"
+
+heatMap$SelectByPlot <- "Reduced dimension plot 1"
+
+# Setting up the annotation function.
+library(org.Hs.eg.db)
+annot.fun <- annotateEnsembl(sce, orgdb=org.Hs.eg.db, keytype="ENSEMBL", rowdata_col="ID")
+
+# Setting up the initial panels.
+init <- DataFrame(Name=c("Reduced dimension plot 1",
+                         "Column data plot 1",
+                         "Feature expression plot 1",
+                         "Row statistics table 1",
+                         "Heat map 1"),
+                  Width=c(4,4,4,8,4))
+
+# Setting up a tour.
+tour <- read.delim("pbmc4k_tour.txt", header=TRUE, sep=";", comment.char="", quote="")
+
+# Running the app.
[…]

### 2018-04-16 manuscript/main.tex
Used 4K PBMC dataset for convenience.

@@
 iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
-We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 68,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells (CITE?).
+We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells (CITE?).
 

### 2018-04-16 tours/tcga_app.R
point to the renamed tour file

@@ -1,7 +1,7 @@
 
 # User inputs ----
 
-tour_steps <- "demo_TCGA.txt"
+tour_steps <- "tcga_tour.txt"
 stopifnot(file.exists(tour_steps))
 
 # Libraries ----

### 2018-04-18 manuscript/main.tex
Fix CS affiliation, add CyTOF ref

@@
 $^{\text{\sf 3}}$Institute for Medical Biostatistics, Epidemiology and Informatics, Mainz, \\
-$^{\text{\sf 4}}$University of Zurich and SIB Swiss Institute of Bioinformatics, and \\
+$^{\text{\sf 4}}$University of Zurich and SIB Swiss Institute of Bioinformatics, Zurich, Switzerland and \\
 $^{\text{\sf 5}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom\\
@@
 iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
-We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells (CITE?).
+We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells \citep{bodenmiller2012cytof}.
 

### 2018-04-24 manuscript/main.tex
checking: citing one of our bibtex entries

--- 2018-04-16 manuscript/main.tex
@@
-\documentclass{bioinfo}
-\copyrightyear{2015} \pubyear{2015}
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Manuscript Category}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% Welcome to Overleaf --- just edit your article on the left,
+% and we'll compile it for you on the right. If you give 
+% someone the link to this page, they can edit at the same
+% time. See the help menu above for more info. Enjoy!
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% For more detailed article preparation guidelines, please see:
+% http://f1000research.com/author-guidelines
+
+\documentclass[10pt,a4paper,twocolumn]{article}
+\usepackage{f1000_styles}
+
+%% Default: numerical citations
+\usepackage[numbers]{natbib}
+
+%% Uncomment this lines for superscript citations instead
+% \usepackage[super]{natbib}
+
+%% Uncomment these lines for author-year citations instead
+% \usepackage[round]{natbib}
+% \let\cite\citep
 
 \begin{document}
-\firstpage{1}
 
-\subtitle{Data and text mining}
-
-\title[iSEE]{iSEE: Interactive SummarizedExperiment Explorer}
-\author[Rue-Albrecht \textit{et~al}.]{Kevin Rue-Albrecht\,$^{\text{\sfb 1,}\dagger}$,
-    Federico Marini\,$^{\text{\sfb 2,3,}\dagger}$,
-    Charlotte Soneson\,$^{\text{\sfb 4,}\dagger}$,
-    and Aaron T. L. Lun\,$^{\text{\sfb 5,}\ast,\dagger}$}
-\address{$^{\text{\sf 1}}$Kennedy Institute of Rheumatology, University of Oxford, Oxford OX3 7FY, United Kingdom, \\
-$^{\text{\sf 2}}$Center for Thrombosis and Hemostasis (CTH), Mainz, \\
-$^{\text{\sf 3}}$Institute for Medical Biostatistics, Epidemiology and Informatics, Mainz, \\
-$^{\text{\sf 4}}$University of Zurich and SIB Swiss Institute of Bioinformatics, and \\
-$^{\text{\sf 5}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom\\
-$^{^\dagger}$These authors contributed equally.}
-
-\corresp{$^\ast$To whom correspondence should be addressed.}
-
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
-
-\editor{Associate Editor: XXXXXXX}
-
-\abstract{\textbf{Summary:} Data exploration is critical to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
-However, most existing tools for interactive visualization are limited to specific assays or analyses.
-Here, we present the iSEE (Interactive SummarizedExperiment Explorer) software package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
-iSEE is directly compatible with many existing R/Bioconductor packages for analyzing high-throughput biological data,
-and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
-We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.\\
-\textbf{Availability:} iSEE is publicly available as an R package from the open-source Bioconductor project (https://bioconductor.org/packages/iSEE).\\
-\textbf{Contact:} \href{aaron.lun@cruk.cam.ac.uk}{aaron.lun@cruk.cam.ac.uk}\\
-\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
-online.}
+\title{\textit{F1000Research} Article Template}
+\titlenote{The title should be detailed enough for someone to know whether the article would be of interest to them, but also concise.
+Please ensure the broadness and claims within the title are appropriate to the content of the article itself.}
+\author[1]{Author Name1}
+\author[2]{Author Name2}
+\affil[1]{Address of author1}
+\affil[2]{Address of author2}
 
 \maketitle
+\thispagestyle{fancy}
 
-\section{Introduction}
-Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics.
-Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
-To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).
+Please list all authors that played a significant role in the research involved in the article.
+Please provide full affiliation information (including full institutional address, ZIP code and e-mail address) for all authors, and identify who is/are the corresponding author(s).
 
-Most existing tools for interactive visualization of biological data are designed for specific assays and analyses, e.g., pRoloc for proteomics \citep{gatto2014mass}, shinyMethyl for methylation \citep{fortin2014shinymethyl}, HTSvis for high-throughput screens \citep{scheeder2017htsvis}.
-Opportunities for customization are generally limited, making it difficult to re-use the same visualization software for new technologies or experimental designs where different aspects of the data are of interest.
-Moreover, standalone tools such as the Loupe Cell Browser from 10X Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}.
-This complicates any coordinated use of these tools with a reproducible, transparent, and statistically rigorous analysis.
+\begin{abstract}
 
-\begin{figure*}[t]
-\includegraphics[width=\textwidth]{pics/Fig1.pdf}
-\caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data.
-New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
-Panel types are available visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
-For these panel types, each sample is represented as a point, which can be coloured according to sample-level metadata or feature-level observations.
-Depending on whether the variables of interest are categorical or numeric, panels will automatically switch between scatter plots, violin plots or ``rectangle plots'' (where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of combination).
-Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
-(B) Information can be transmitted between panels according to a user-specified scheme.
-Here, the selection of feature $X$ in the row statistics table determines the y-axis of the feature expression plot, and colours the samples in the reduced dimension plot by the expression of $X$.
-Selection of points in the reduced dimension plot (dotted blue line) also determines the samples that are shown in the column data (i.e., sample metadata) plot;
-further selection of points in the column data plot determines the samples that are shown in the heatmap.
-}
-\label{fig:iSEE}
-\end{figure*}
+Abstracts should be up to 300 words and provide a succinct summary of the article.
+Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasize the importance of the work described in the article.
+Citations should not be used in the abstract, and the use of abbreviations should be minimized.
+If you are writing a Research or Systematic Review article, please structure your abstract into Background, Methods, Results, and Conclusions.
 
-\section{Description of the iSEE interface}
-Here, we present the iSEE software package for interactive data exploration.
-iSEE is implemented in R using the shiny framework \citep{chang2017shiny} and exploits data architectures from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
-The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
-This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively.
-Per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results.
-All of these data aspects can be directly examined in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
-The flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
-By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in R/Bioconductor packages.
-%In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).
+\end{abstract}
 
-A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
-Users can define arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting tables and plots highlight the corresponding data points in receiving panels.
-This facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to one or more feature expression plots.
-The linking paradigm extends to multiple plots whereby a plot can transmit to multiple receivers, or a receiving plot can itself transmit to another plot.
-This allows users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}, with the advantage of being extensible to any (meta)data present in a SummarizedExperiment object.
-Information can similarly be transferred between tables and plots, e.g., selecting a row determines how points are coloured in receiving plots.
+\section*{Keywords}
 
-iSEE also offers a host of other customizable features to improve data exploration.
-Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each metadata variable and assay type across all panels in the application.
-Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez.
-% do we need to cite them?
-at least ENSEMBL has updated citable papers
-For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
-After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
-iSEE also automatically memorizes the exact R code that was used to generate every plot and the existing links,
-extending previous work by \cite{marini2016interrepro}.
-This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.
+Please list up to eight keywords to help readers interested in your article find it more easily.
 
-\section{Conclusion}
-iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
-We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells (CITE?).
+\clearpage
 
-\section*{Acknowledgements}
-We thank the organizers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
-We also thank members of the Bioconductor community for their helpful suggestions.
-\vspace*{-12pt}
+\section*{Introduction}
 
-\section*{Funding}
-ATLL was supported by core funding from Cancer Research UK (award no. 17197 to Dr.\ John Marioni).
-The work of FM is supported by the German Federal Ministry of Education and Research (BMBF 01EO1003).
-\vspace*{-12pt}
+The format of the main body of the article is flexible: it should be concise and in the format most appropriate to displaying the content of the article.
 
-%\bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
-\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
-%\bibliographystyle{bioinformatics}
-%
-%\bibliographystyle{plain}
-%
-\bibliography{ref}
+Some examples of commonly used \LaTeX{}  commands and features are listed below, to help you get started.
+
+\subsection*{Sections}
+
+Use section and subsection commands to organize your document. \LaTeX{} handles all the formatting and numbering automatically. Use ref and label commands for cross-references.
+
+\subsection*{Tables}
+
+Use the table and tabledata commands for basic tables --- see Table~\ref{tab:widgets}, for example.
+\begin{table}[h!]
+\hrule \vspace{0.1cm}
+\caption{\label{tab:widgets}An example of a simple table with caption.}
+\centering
+\begin{tabledata}{llr} 
+\header First name & Last Name & Grade \\ 
+\row John & Doe & $7.5$ \\ 
+\row Richard & Miles & $2$ \\ 
+\end{tabledata}
+\end{table}
+
+\subsection*{Figures}
+You can upload a figure (JPEG, PNG or PDF) using the files menu. To include it in your document, use the includegraphics command (see the example below in the source code).
+
+Please give figures appropriate filenames eg: figure1.pdf, figure2.png.
+
+Figure legends should briefly describe the key messages of the figure such that the figure can stand alone from the main text.
+However, all figures should also be discussed in the article text.
+Each legend should have a concise title of no more than 15 words.
+The legend itself should be succinct, while still explaining all symbols and abbreviations.
+Avoid lengthy descriptions of methods.
+
+For any figures reproduced from another publication (as long as appropriate permission has been obtained from the copyright holder —see under the heading 'Submission'), please include a line in the legend to state that: 'This figure has been reproduced with kind permission from [include original publication citation]'.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.4\textwidth]{frog.jpg}
+\caption{\label{fig:your-figure}Your figure legend goes here; it should be succinct, while still explaining all symbols and abbreviations. }
+\end{figure}
+
+\subsection*{Mathematics}
+
+\LaTeX{} is great at typesetting mathematics.
+Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
+$$S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
+      = \frac{1}{n}\sum_{i}^{n} X_i$$
+denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.
+
+\subsection*{Lists}
+
+You can make lists with automatic numbering \dots
+
+\begin{enumerate}
+\item Like this,
+\item and like this.
+\end{enumerate}
+\dots or bullet points \dots
+\begin{itemize}
+\item Like this,
+\item and like this.
+\end{itemize}
+
+\section*{Methods}
+Methods should include a brief discussion of allowances made (if any) for controlling bias or unwanted sources of variability, and the limitations of the datasets.
+
+\section*{Results}
+This section is not essential for Web Tool papers.
+
+\section*{Discussion}
+The discussion should include the implications of the article results in view of prior work in this field.
+
+\section*{Conclusions}
+Please state what you think are the main conclusions that can be realistically drawn from the findings in the paper, taking care not to make claims that cannot be supported.
+
+\subsection*{Author contributions}
+In order to give appropriate credit to each author of an article, the individual
+contributions of each author to the manuscript should be detailed in this section. We
+recommend using author initials and then stating briefly how they contributed.
+
+\subsection*{Competing interests}
+All financial, personal, or professional competing interests for any of the authors that
+could be construed to unduly influence the content of the article must be disclosed and
+will be displayed alongside the article.
+
+\subsection*{Grant information}
+Please state who funded the work discussed in this article, whether it is your employer,
+a grant funder etc. Please do not list funding that you have that is not relevant to this
+specific piece of research. For each funder, please state the funder’s name, the grant
+number where applicable, and the individual to whom the grant was assigned.
+If your work was not funded by any grants, please include the line: ‘The author(s)
+declared that no grants were involved in supporting this work.’
+
+\subsection*{Acknowledgements}
+This section should acknowledge anyone who contributed to the research or the
+article but who does not qualify as an author based on the criteria provided earlier
+(e.g. someone or an organisation that provided writing assistance). Please state how
+they contributed; authors should obtain permission to acknowledge from all those
+mentioned in the Acknowledgements section.
+
+Please do not list grant funding in this section.
+
+{\small\bibliographystyle{unsrtnat}
+\bibliography{ref}}
+
+\bigskip
+References can be listed in any standard referencing style that uses a numbering system
+(i.e. not Harvard referencing style), and should be consistent between references within
+a given article.
+
+Reference management systems such as Zotero provide options for exporting bibliographies as Bib\TeX{} files.
+Bib\TeX{} is a bibliographic tool that is used with \LaTeX{} to help organize the user's references and create a bibliography.
+This template contains an example of such a file, \texttt{sample.bib}, which can be replaced with your own.
+Use the \verb|\cite| command  to create in-text citations, \cite{zheng2017massively}
+%like this \cite{Smith:2012qr} and this \cite{Smith:2013jd}.
+
+% See this guide for more information on BibTeX:
+% http://libguides.mit.edu/content.php?pid=55482&sid=406343
+
+% For more author guidance please see:
+% http://f1000research.com/author-guidelines
+
+% When all authors are happy with the paper, use the 
+% ‘Submit to F1000Research' button from the menu above
+% to submit directly to the open life science journal F1000Research.
+
+% Please note that this template results in a draft pre-submission PDF document.
+% Articles will be professionally typeset when accepted for publication.
+
+% We hope you find the F1000Research Overleaf template useful,
+% please let us know if you have any feedback using the help menu above.
 

### 2018-04-24 manuscript/main.tex
author list is full. Todo: dagger or similar for equal contribution

@@
 
-\title{\textit{F1000Research} Article Template}
+\title{iSEE: Interactive SummarizedExperiment Explorer}
 \titlenote{The title should be detailed enough for someone to know whether the article would be of interest to them, but also concise.
 Please ensure the broadness and claims within the title are appropriate to the content of the article itself.}
-\author[1]{Author Name1}
-\author[2]{Author Name2}
-\affil[1]{Address of author1}
-\affil[2]{Address of author2}
+\author[1]{Kevin Rue-Albrecht}
+\author[2,3]{Federico Marini}
+\author[4]{Charlotte Soneson}
+\author[5]{Aaron T. L. Lun}
+
+\affil[1]{Kennedy Institute of Rheumatology, University of Oxford, Oxford OX3 7FY, United Kingdom}
+\affil[2]{Center for Thrombosis and Hemostasis (CTH), University Medical Center of the Johannes Gutenberg University Mainz, Mainz}
+\affil[3]{Institute for Medical Biostatistics, Epidemiology and Informatics, University Medical Center of the Johannes Gutenberg University Mainz, Mainz}
+\affil[4]{University of Zurich and SIB Swiss Institute of Bioinformatics, Zurich, Switzerland}
+\affil[5]{Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom}
+
+% TODO: $^{^\dagger}$These authors contributed equally.}
 

### 2018-04-24 manuscript/main.tex
first draft of ban format, ported entirely to f1000res format

--- 2018-04-18 manuscript/main.tex
@@
-\documentclass{bioinfo}
-\copyrightyear{2015} \pubyear{2015}
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Manuscript Category}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% Welcome to Overleaf --- just edit your article on the left,
+% and we'll compile it for you on the right. If you give 
+% someone the link to this page, they can edit at the same
+% time. See the help menu above for more info. Enjoy!
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% For more detailed article preparation guidelines, please see:
+% http://f1000research.com/author-guidelines
+
+\documentclass[10pt,a4paper,twocolumn]{article}
+\usepackage{f1000_styles}
+
+%% Default: numerical citations
+\usepackage[numbers]{natbib}
+
+%% Uncomment this lines for superscript citations instead
+% \usepackage[super]{natbib}
+
+%% Uncomment these lines for author-year citations instead
+% \usepackage[round]{natbib}
+% \let\cite\citep
 
 \begin{document}
-\firstpage{1}
 
-\subtitle{Data and text mining}
+\title{iSEE: Interactive SummarizedExperiment Explorer}
+\titlenote{The title should be detailed enough for someone to know whether the article would be of interest to them, but also concise.
+Please ensure the broadness and claims within the title are appropriate to the content of the article itself.}
+\author[1]{Kevin Rue-Albrecht}
+\author[2,3]{Federico Marini}
+\author[4]{Charlotte Soneson}
+\author[5]{Aaron T. L. Lun}
 
-\title[iSEE]{iSEE: Interactive SummarizedExperiment Explorer}
-\author[Rue-Albrecht \textit{et~al}.]{Kevin Rue-Albrecht\,$^{\text{\sfb 1,}\dagger}$,
-    Federico Marini\,$^{\text{\sfb 2,3,}\dagger}$,
-    Charlotte Soneson\,$^{\text{\sfb 4,}\dagger}$,
-    and Aaron T. L. Lun\,$^{\text{\sfb 5,}\ast,\dagger}$}
-\address{$^{\text{\sf 1}}$Kennedy Institute of Rheumatology, University of Oxford, Oxford OX3 7FY, United Kingdom, \\
-$^{\text{\sf 2}}$Center for Thrombosis and Hemostasis (CTH), Mainz, \\
-$^{\text{\sf 3}}$Institute for Medical Biostatistics, Epidemiology and Informatics, Mainz, \\
-$^{\text{\sf 4}}$University of Zurich and SIB Swiss Institute of Bioinformatics, Zurich, Switzerland and \\
-$^{\text{\sf 5}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom\\
-$^{^\dagger}$These authors contributed equally.}
+\affil[1]{Kennedy Institute of Rheumatology, University of Oxford, Oxford OX3 7FY, United Kingdom}
+\affil[2]{Center for Thrombosis and Hemostasis (CTH), University Medical Center of the Johannes Gutenberg University Mainz, Mainz}
+\affil[3]{Institute for Medical Biostatistics, Epidemiology and Informatics, University Medical Center of the Johannes Gutenberg University Mainz, Mainz}
+\affil[4]{University of Zurich and SIB Swiss Institute of Bioinformatics, Zurich, Switzerland}
+\affil[5]{Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom}
 
-\corresp{$^\ast$To whom correspondence should be addressed.}
+% TODO: $\dagger$ These authors contributed equally.}
+% TODO: symbol for corresponding author e.g. $\ast$
 
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+\maketitle
+\thispagestyle{fancy}
 
-\editor{Associate Editor: XXXXXXX}
+Please list all authors that played a significant role in the research involved in the article.
+Please provide full affiliation information (including full institutional address, ZIP code and e-mail address) for all authors, and identify who is/are the corresponding author(s).
 
-\abstract{\textbf{Summary:} Data exploration is critical to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
+% TODO: full address and emails too?
+
+\begin{abstract}
+
+\textbf{Summary:} Data exploration is critical to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
 However, most existing tools for interactive visualization are limited to specific assays or analyses.
@@
 \textbf{Availability:} iSEE is publicly available as an R package from the open-source Bioconductor project (https://bioconductor.org/packages/iSEE).\\
-\textbf{Contact:} \href{aaron.lun@cruk.cam.ac.uk}{aaron.lun@cruk.cam.ac.uk}\\
-\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
-online.}
+\textbf{Contact:} aaron.lun@cruk.cam.ac.uk\\
+\textbf{Supplementary information:} Supplementary data are available online.
 
-\maketitle
+% Abstracts should be up to 300 words and provide a succinct summary of the article.
+Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasize the importance of the work described in the article.
+Citations should not be used in the abstract, and the use of abbreviations should be minimized.
+If you are writing a Research or Systematic Review article, please structure your abstract into Background, Methods, Results, and Conclusions.
 
-\section{Introduction}
+\end{abstract}
+
+\section*{Keywords}
+
+Please list up to eight keywords to help readers interested in your article find it more easily.
+
+% TODO: pick up nice keywords!
+
+\clearpage
+
+\section*{Introduction}
 Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics.
@@
 
-\section{Description of the iSEE interface}
+\section*{Description of the iSEE interface}
 Here, we present the iSEE software package for interactive data exploration.
@@
 
-\section{Conclusion}
+\section*{Conclusion}
 iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
@@
 
-\section*{Acknowledgements}
+\subsection*{Author contributions}
+In order to give appropriate credit to each author of an article, the individual
+contributions of each author to the manuscript should be detailed in this section. We
+recommend using author initials and then stating briefly how they contributed.
+
+\subsection*{Competing interests}
+All financial, personal, or professional competing interests for any of the authors that
+could be construed to unduly influence the content of the article must be disclosed and
+will be displayed alongside the article.
+
+\subsection*{Grant information}
+ATLL was supported by core funding from Cancer Research UK (award no. 17197 to Dr.\ John Marioni).
+The work of FM is supported by the German Federal Ministry of Education and Research (BMBF 01EO1003).
+
+\subsection*{Acknowledgements}
 We thank the organizers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
 We also thank members of the Bioconductor community for their helpful suggestions.
-\vspace*{-12pt}
 
-\section*{Funding}
-ATLL was supported by core funding from Cancer Research UK (award no. 17197 to Dr.\ John Marioni).
-The work of FM is supported by the German Federal Ministry of Education and Research (BMBF 01EO1003).
-\vspace*{-12pt}
+{\small\bibliographystyle{unsrtnat}
+\bibliography{ref}}
 
-%\bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
-\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
-%\bibliographystyle{bioinformatics}
-%
-%\bibliographystyle{plain}
-%
-\bibliography{ref}
+\bigskip
+References can be listed in any standard referencing style that uses a numbering system
+(i.e. not Harvard referencing style), and should be consistent between references within
+a given article.
+
+Reference management systems such as Zotero provide options for exporting bibliographies as Bib\TeX{} files.
+Bib\TeX{} is a bibliographic tool that is used with \LaTeX{} to help organize the user's references and create a bibliography.
+This template contains an example of such a file, \texttt{sample.bib}, which can be replaced with your own.
+Use the \verb|\cite| command  to create in-text citations, \cite{zheng2017massively}
+%like this \cite{Smith:2012qr} and this \cite{Smith:2013jd}.
+
+% See this guide for more information on BibTeX:
+% http://libguides.mit.edu/content.php?pid=55482&sid=406343
+
+% For more author guidance please see:
+% http://f1000research.com/author-guidelines
+
+% When all authors are happy with the paper, use the 
+% ‘Submit to F1000Research' button from the menu above
+% to submit directly to the open life science journal F1000Research.
+
+% Please note that this template results in a draft pre-submission PDF document.
+% Articles will be professionally typeset when accepted for publication.
+
+% We hope you find the F1000Research Overleaf template useful,
+% please let us know if you have any feedback using the help menu above.
 

### 2018-04-24 manuscript/main.tex
fixed equal contribution plus corresponding author symbols
IMBEI as acronym added

@@
 Please ensure the broadness and claims within the title are appropriate to the content of the article itself.}
-\author[1]{Kevin Rue-Albrecht}
-\author[2,3]{Federico Marini}
-\author[4]{Charlotte Soneson}
-\author[5]{Aaron T. L. Lun}
+\author[1,$\dagger$]{Kevin Rue-Albrecht}
+\author[2,3,$\dagger$]{Federico Marini}
+\author[4,$\dagger$]{Charlotte Soneson}
+\author[5,$\dagger$,$\ast$]{Aaron T. L. Lun}
 
@@
 \affil[2]{Center for Thrombosis and Hemostasis (CTH), University Medical Center of the Johannes Gutenberg University Mainz, Mainz}
-\affil[3]{Institute for Medical Biostatistics, Epidemiology and Informatics, University Medical Center of the Johannes Gutenberg University Mainz, Mainz}
+\affil[3]{Institute for Medical Biostatistics, Epidemiology and Informatics (IMBEI), University Medical Center of the Johannes Gutenberg University Mainz, Mainz}
 \affil[4]{University of Zurich and SIB Swiss Institute of Bioinformatics, Zurich, Switzerland}
 \affil[5]{Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom}
-
-% TODO: $\dagger$ These authors contributed equally.}
-% TODO: symbol for corresponding author e.g. $\ast$
+\affil[$\dagger$]{These authors contributed equally.}
+\affil[$\ast$]{To whom correspondence should be addressed.}
 

### 2018-04-24 manuscript/main.tex
removed all the default content about bibtex, unneeded

@@
 
-Reference management systems such as Zotero provide options for exporting bibliographies as Bib\TeX{} files.
-Bib\TeX{} is a bibliographic tool that is used with \LaTeX{} to help organize the user's references and create a bibliography.
-This template contains an example of such a file, \texttt{sample.bib}, which can be replaced with your own.
-Use the \verb|\cite| command  to create in-text citations, \cite{zheng2017massively}
-%like this \cite{Smith:2012qr} and this \cite{Smith:2013jd}.
-
-% See this guide for more information on BibTeX:
-% http://libguides.mit.edu/content.php?pid=55482&sid=406343
-
 % For more author guidance please see:

### 2018-04-26 manuscript/main.tex
Prepared sections to flesh out, and minor edits here and there.

@@
 
-Please list all authors that played a significant role in the research involved in the article.
+% Please list all authors that played a significant role in the research involved in the article.
 Please provide full affiliation information (including full institutional address, ZIP code and e-mail address) for all authors, and identify who is/are the corresponding author(s).
-
-% TODO: full address and emails too?
+% TODO: full address and emails too? if so, in which format do we provide them?
 
@@
 
-Please list up to eight keywords to help readers interested in your article find it more easily.
+% Please list up to eight keywords to help readers interested in your article find it more easily.
 
-% TODO: pick up nice keywords!
+Single-cell, gene expression, transcriptomics, proteomics, visualization, interactive, shiny, Bioconductor
+
+% TODO: refine keywords
 
@@
 Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
-To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).
+To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions, microarray probe sets).
 
@@
 
-\section*{Description of the iSEE interface}
+\section*{iSEE supports multiple genomics data types}
 Here, we present the iSEE software package for interactive data exploration.
@@
 The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
-This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively.
-Per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results.
-All of these data aspects can be directly examined in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
-The flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
-By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in R/Bioconductor packages.
+The SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively.
+For instance, these may represent gene expression matrices of raw counts along with different normalization strategies.
+In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; those may include experimental metadata as well as any analysis results.
+All of these data aspects can be readily examined alongside one another in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
+
+Notably, the flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
+As a consequence, the widespread adoption of the SummarizedExperiment class and its specialized extensions such as the SingleCellExperiment class enables iSEE to offer immediate interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in R/Bioconductor packages.
 %In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).
+% KRA: any other child class like SingleCellExperiment worth mentioning? e.g. the BSseq class
+
+\section*{Description of the iSEE interface}
+
+In this section, give a brief description of each panel type.
+
+% the previous section mainly focuses on the input object
+
+\section*{Dynamic linking between panels}
 
@@
 Users can define arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting tables and plots highlight the corresponding data points in receiving panels.
-This facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to one or more feature expression plots.
-The linking paradigm extends to multiple plots whereby a plot can transmit to multiple receivers, or a receiving plot can itself transmit to another plot.
+This feature facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to one or more feature expression plots.
+The linking paradigm extends to multiple plots whereby a plot can transmit to multiple receivers, or a receiving plot can itself transmit its own selection to another plot.
 This allows users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}, with the advantage of being extensible to any (meta)data present in a SummarizedExperiment object.
@@
 
+\section*{Advanced features and customization}
+
+% KRA: Also, we should probably highlight somewhere that additional inputs can customize the startup state of the application
+
+\subsection*{ExperimentColorMap: custom color maps}
+
 iSEE also offers a host of other customizable features to improve data exploration.
 Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each metadata variable and assay type across all panels in the application.
+% TODO: KRA develop here
+
+\subsection*{Linked feature annotations}
+
 Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez.
@@
 at least ENSEMBL has updated citable papers
+
+\subsection*{Density-dependent downsampling for speed}
+
 For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
+
+\subsection*{Code tracking and reproducibility}
+
 After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
@@
 
+\section*{Demonstration data sets and tours}
+
+\subsection*{Single-cell plate-based RNA-seq}
+% TODO: better title, or do we drop this allen section altogether?
+
+We demonstrate the application of iSEE to examine a subset of 379 cells from of mouse visual cortex \citep{tasic2016allen}.
+
+% TODO: embed the allen demo data set
+
+\subsection*{Single-cell droplet-based RNA-seq}
+% TODO: better title, or do we drop the allen section altogether (above)?
+
+We also demonstrate the use of iSEE to explore a larger data set of 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively}.
+
+% embed the allen demo data set
+
+\subsection*{Bulk RNA-seq}
+
+We also demonstrate how iSEE seamlessly scales up with larger data sets as the TCGA RNA-seq collection of measurements for 23,368 genes across 7,706 tumor samples \citep{piccolo2015TCGA}.
+
+% TODO: embed the TCGA demo data set
+
+\subsection*{CyTOF}
+
+Lastly, we also showcase the use of iSEE to vizualise and refine gating a gating analysis, followed by the investigation of a functional marker between two sample groups, using a published mass cytometry (CyTOF) data set \citep{bodenmiller2012cytof}.
+
+% TODO: embed the TCGA demo data set
+
 \section*{Conclusion}
@@
 We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells \citep{bodenmiller2012cytof}.
+% TODO: repurpose the conclusion if we showcase the examples in earlier sections
 

### 2018-04-27 manuscript/main.tex
added some flesh: description of ui, functionality, and expanded repro+tours

@@
 %% Default: numerical citations
-\usepackage[numbers]{natbib}
+% \usepackage[numbers]{natbib}
 
@@
 %% Uncomment these lines for author-year citations instead
-% \usepackage[round]{natbib}
-% \let\cite\citep
+\usepackage[round]{natbib}
+\let\cite\citep
 
@@
 
-\section*{iSEE supports multiple genomics data types}
+\section*{Implementation}
 Here, we present the iSEE software package for interactive data exploration.
@@
 
-\section*{Description of the iSEE interface}
+% \section*{Description of the iSEE interface} % FM: if required, "turn on the section"?
+The layout of the iSEE user interface is built on top of the shinydashboard package. 
+The header contains three dropdown menus, related to internal iSEE diagnostics function, documentation, and additional information.
+% TODO, FM: maybe we can/should cross-reference to the upcoming sections?
+The sidebar of iSEE contains a dropdown menu and a button for creating new plots or tables (referred to as ``panels'') in the interface.
+Panels can be added, canceled, adjusted in width and height, and re-organized, by clicking on the corresponding color-coded tabs.
+The main element in the body of iSEE is the combination of panels, generated (and optionally linked to one another) according to the users' actions.  
+Besides the package documentation and vignette, users can get familiar with the main elements of the user interface by taking an interactive tour (based on the rintrojs package []ganz[]), which will allow them to learn the basic usage mechanisms by doing.
 
-In this section, give a brief description of each panel type.
+\section*{Functionality}
+Currently, there are currently six different panel types that can be generated with iSEE.
+For each panel, three different set of parameters are available in collapsible boxes.
+Data parameters control settings specific to each type of plot; Visual parameters specify parameters that will determine the aspect of the plot (coloring, point features, ...); Selection parameters relate to the point selections and link relationships to other plots.
 
-% the previous section mainly focuses on the input object
+\begin{itemize}
+\item Reduced dimension plots -  Display any reduced dimension results (e.g., PCA, t-SNE), extracted from the reducedDim slot of a SingleCellExperiment object is supplied to iSEE, enabling efficient exploration of high-dimensional datasets.
+\item Column data plots - Visualize sample metadata stored in the SummarizedExperiment object, whereas plots can assume various forms, depending on the nature of the data on the x- and y-axes.
+\item Feature expression plots - Visualize the expression values for a particular feature (e.g., a gene) across the samples on the y-axis, often resulting in (grouped) violin plots or scatter plots, if the X axis is set to categorical or continuous variables, respectively.
+\item Row statistics tables - Present the values of the rowData slot for the SummarizedExperiment object.
+A typical usage for such tables is to link to other plots, in order to determine the features to use for plotting, or coloring.
+\item Row data plots - Visualize the information stored in the rowData slot of a SummarizedExperiment object, in a behavior that follows the implementation of column data plots.
+\item Heat maps - Visualize overviews of the data at hand, in the form of color-coded expression matrices, where features such as genes are the rows, and samples are the columns.
+Selection of features is possible from dedicated widgets, which also allow subsetting based on other panels.
+\end{itemize}
 
-\section*{Dynamic linking between panels}
+Coloring of points (i.e., samples) on each plot can be achieved in different ways, where the default is no color scheme and results in black data points. 
+Any column from the metadata of the SummarizedExperiment can be used, and the same is true for the expression values of a particular feature.
+The plotting routines automatically adjust the scale to use, based on the chosen variable.
 
+\subsection*{Dynamic linking between panels}
 A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
@@
 
-\section*{Advanced features and customization}
+\subsection*{Advanced features and customization}
 
-% KRA: Also, we should probably highlight somewhere that additional inputs can customize the startup state of the application
+Additional parameters, provided when launching the iSEE application, can customize the startup state of the application
+The number and identity of the panels, as well as their inter-relationships can also be specified at initialization, thus avoiding the need to click through the choices to obtain the desired panel setup.
+An example of such customization is provided in %% TODO %%
+Users can refer to the documentation of ?defaults for the full list of values that can be specified in iSEE.
 
@@
 
+iSEE also automatically memorizes the exact R code that was used to generate every plot and the existing links, extending previous work by \cite{marini2016interrepro}.
+A modal popup window, with a shinyAce-based text editor, where the code is formatted and displayed with syntax highlighting, is opened when clicking on the `Extract the R code'' button from the diagnostics dropdown menu in the header.
+This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.
+Additionally, the ``Examine panel chart'' functionality, in the iSEE diagnostics dropdown menu, returns a graph representation of the existing links and point selection among the open panels.
+This can be very useful in sessions that include a large number of panels.
+Clicking on the ``Display panel settings'' button from the same dropdown menu returns the required lines of code that underlie the current setup of currently open panels.
+
+\subsection*{App tours and deployment}
 After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
-iSEE also automatically memorizes the exact R code that was used to generate every plot and the existing links,
-extending previous work by \cite{marini2016interrepro}.
-This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.
+This normally requires the setup of a Shiny Server, % TODO: FM, do we want to provide some basic instructions on how to do it?, 
+as well as the preparation of the dataset at hand in form of a SummarizedExperiment - which many workflows available in Bioconductor adopt as main representation.
+Tours can be provided to the iSEE function in the form of simple data frames, containing the list of DOM elements in the Shiny app to anchor the corresponding descriptions. 
+Examples of such tours can be found in the inst/extdata folder of the iSEE package, as well as in the GitHub repository of this manuscript, https://github.com/LTLA/iSEE2018/tree/master/tours.
 
-\section*{Demonstration data sets and tours}
+%\section*{Demonstration data sets and tours}
+\section*{Use cases} % FM: used often in other software articles in F1000res
+
+To demonstrate the functionality and flexibility of the iSEE package, we applied our method to a number of publicly available datasets.
+Details are provided in the tour subfolder of the GitHub repository for this manuscript (https://github.com/LTLA/iSEE2018), and for each dataset a combination of _app.R and _tour.txt is provided.
+An instance of iSEE running on the respective datasets is provided as a stand-alone webpages, for demonstration purposes without the need of software installation.
 
@@
 % TODO: embed the allen demo data set
+http://shiny.imbei.uni-mainz.de:3838/iSEE/
 
@@
 % embed the allen demo data set
+http://shiny.imbei.uni-mainz.de:3838/iSEE_PBMC4k/
 
-\subsection*{Bulk RNA-seq}
+\subsection*{Bulk RNA-seq from TCGA}
 
@@
 % TODO: embed the TCGA demo data set
+http://shiny.imbei.uni-mainz.de:3838/iSEE_TCGA/
 
@@
 % TODO: embed the TCGA demo data set
+http://shiny.imbei.uni-mainz.de:3838/iSEE_cytof/
 
@@
 % TODO: repurpose the conclusion if we showcase the examples in earlier sections
+% TODO: sell the idea that for every dataset it is possible to explore and showcase feats of interest
 
@@
 recommend using author initials and then stating briefly how they contributed.
+
+\subsection*{Data and software availability}
+- source code
+- archived source
+- package at...
+- shiny at...
+- docker/singularity containers
+- conda pkg
+- Software license: MIT
 

### 2018-04-27 manuscript/main.tex
further flesh with some more info, plus cleaning up template leftovers

@@
 \section*{Use cases} % FM: used often in other software articles in F1000res
-
 To demonstrate the functionality and flexibility of the iSEE package, we applied our method to a number of publicly available datasets.
@@
 % TODO: better title, or do we drop this allen section altogether?
+We demonstrate the application of iSEE to examine a subset of 379 cells from of mouse visual cortex \citep{tasic2016allen}.
+This example can be reproduced also by calling
 
-We demonstrate the application of iSEE to examine a subset of 379 cells from of mouse visual cortex \citep{tasic2016allen}.
+library(iSEE)
+example(iSEE, ask = F)
 
@@
 % TODO: better title, or do we drop the allen section altogether (above)?
-
 We also demonstrate the use of iSEE to explore a larger data set of 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively}.
-
+% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
 % embed the allen demo data set
@@
 \subsection*{Bulk RNA-seq from TCGA}
-
 We also demonstrate how iSEE seamlessly scales up with larger data sets as the TCGA RNA-seq collection of measurements for 23,368 genes across 7,706 tumor samples \citep{piccolo2015TCGA}.
-
+% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
 % TODO: embed the TCGA demo data set
@@
 \subsection*{CyTOF}
+Lastly, we also showcase the use of iSEE to vizualise and refine gating a gating analysis, followed by the investigation of a functional marker between two sample groups, using a published mass cytometry (CyTOF) data set \citep{bodenmiller2012cytof}.
+% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
 
-Lastly, we also showcase the use of iSEE to vizualise and refine gating a gating analysis, followed by the investigation of a functional marker between two sample groups, using a published mass cytometry (CyTOF) data set \citep{bodenmiller2012cytof}.
-
-% TODO: embed the TCGA demo data set
 http://shiny.imbei.uni-mainz.de:3838/iSEE_cytof/
@@
 % TODO: repurpose the conclusion if we showcase the examples in earlier sections
-% TODO: sell the idea that for every dataset it is possible to explore and showcase feats of interest
+% TODO: sell the idea that for every dataset it is possible to explore and showcase feats of interest. Something like...
+We expect the functionality and flexibility of the iSEE package to be essential when exploring and showcasing relevant features of any dataset which can be represented in the form of a SummarizedExperiment object, in such a way that a public instance of iSEE can ideally accompany any publication, e.g.
+reproducing the manuscript figures, but also allowing any user to foster hypothesis generation with efficient interactive and reproducible data analysis.
 
@@
 \subsection*{Data and software availability}
-- source code
-- archived source
-- package at...
-- shiny at...
-- docker/singularity containers
-- conda pkg
-- Software license: MIT
+The iSEE package is available at https://bioconductor.org/packages/iSEE
+
+Source code of the development version is available at https://github.com/csoneson/iSEE
+
+Archived source code as at the time of publication is available at % TODO
+
+An instance of the package running on a public Shiny Server is available at http://shiny.imbei.uni-mainz.de:3838/iSEE/
+
+% TODO: docker/singularity containers
+
+% TODO:conda pkg
+% The iSEE package is also distributed via the bioconda channel for the conda package manager at...https://anaconda.org/bioconda/bioconductor-isee % still needs to be finalized!!
+
+Software license: MIT
 
@@
 \bigskip
-References can be listed in any standard referencing style that uses a numbering system
-(i.e. not Harvard referencing style), and should be consistent between references within
-a given article.
+% References can be listed in any standard referencing style that uses a numbering system
+% (i.e. not Harvard referencing style), and should be consistent between references within
+% a given article.
 

### 2018-04-27 manuscript/main.tex
l'URLo di munch - adding url where required

@@
 \usepackage{f1000_styles}
+\usepackage{url}
 
@@
 We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.\\
-\textbf{Availability:} iSEE is publicly available as an R package from the open-source Bioconductor project (https://bioconductor.org/packages/iSEE).\\
+\textbf{Availability:} iSEE is publicly available as an R package from the open-source Bioconductor project (\url{https://bioconductor.org/packages/iSEE}).\\
 \textbf{Contact:} aaron.lun@cruk.cam.ac.uk\\
@@
 Tours can be provided to the iSEE function in the form of simple data frames, containing the list of DOM elements in the Shiny app to anchor the corresponding descriptions. 
-Examples of such tours can be found in the inst/extdata folder of the iSEE package, as well as in the GitHub repository of this manuscript, https://github.com/LTLA/iSEE2018/tree/master/tours.
+Examples of such tours can be found in the inst/extdata folder of the iSEE package, as well as in the GitHub repository of this manuscript, \url{https://github.com/LTLA/iSEE2018/tree/master/tours}.
 
@@
 To demonstrate the functionality and flexibility of the iSEE package, we applied our method to a number of publicly available datasets.
-Details are provided in the tour subfolder of the GitHub repository for this manuscript (https://github.com/LTLA/iSEE2018), and for each dataset a combination of _app.R and _tour.txt is provided.
+Details are provided in the tour subfolder of the GitHub repository for this manuscript (\url{https://github.com/LTLA/iSEE2018}), and for each dataset a combination of \_app.R and \_tour.txt is provided.
 An instance of iSEE running on the respective datasets is provided as a stand-alone webpages, for demonstration purposes without the need of software installation.
@@
 % TODO: embed the allen demo data set
-http://shiny.imbei.uni-mainz.de:3838/iSEE/
+\url{http://shiny.imbei.uni-mainz.de:3838/iSEE/}
 
@@
 % embed the allen demo data set
-http://shiny.imbei.uni-mainz.de:3838/iSEE_PBMC4k/
+\url{http://shiny.imbei.uni-mainz.de:3838/iSEE_PBMC4k/}
 
@@
 % TODO: embed the TCGA demo data set
-http://shiny.imbei.uni-mainz.de:3838/iSEE_TCGA/
+\url{http://shiny.imbei.uni-mainz.de:3838/iSEE_TCGA/}
 
@@
 
-http://shiny.imbei.uni-mainz.de:3838/iSEE_cytof/
+\url{http://shiny.imbei.uni-mainz.de:3838/iSEE_cytof/}
 
@@
 iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
-We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells \citep{bodenmiller2012cytof}.
+We provide a number of demonstrations at \url{https://github.com/LTLA/iSEE2018}, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells \citep{bodenmiller2012cytof}.
 % TODO: repurpose the conclusion if we showcase the examples in earlier sections
@@
 \subsection*{Data and software availability}
-The iSEE package is available at https://bioconductor.org/packages/iSEE
+The iSEE package is available at \url{https://bioconductor.org/packages/iSEE}
 
-Source code of the development version is available at https://github.com/csoneson/iSEE
+Source code of the development version is available at \url{https://github.com/csoneson/iSEE}
 
@@
 
-An instance of the package running on a public Shiny Server is available at http://shiny.imbei.uni-mainz.de:3838/iSEE/
+An instance of the package running on a public Shiny Server is available at \url{http://shiny.imbei.uni-mainz.de:3838/iSEE/}
 

### 2018-04-28 manuscript/main.tex
more tweaks; added reference to shinydashboard

@@
 New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
-Panel types are available visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
+Panel types are available to visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
 For these panel types, each sample is represented as a point, which can be coloured according to sample-level metadata or feature-level observations.
@@
 Here, we present the iSEE software package for interactive data exploration.
-iSEE is implemented in R using the shiny framework \citep{chang2017shiny} and exploits data architectures from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
+iSEE is implemented in R using the Shiny framework \citep{chang2017shiny} and exploits data architectures from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
 The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
-The SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively.
-For instance, these may represent gene expression matrices of raw counts along with different normalization strategies.
+The SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where rows and columns represent genomic features and biological samples, respectively.
+For instance, individual assays may represent gene expression matrices of raw counts along with different normalization strategies.
 In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; those may include experimental metadata as well as any analysis results.
@@
 Notably, the flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
-As a consequence, the widespread adoption of the SummarizedExperiment class and its specialized extensions such as the SingleCellExperiment class enables iSEE to offer immediate interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in R/Bioconductor packages.
+As a consequence, the widespread adoption of the SummarizedExperiment class and its specialized extensions such as the SingleCellExperiment class enables iSEE to offer immediate interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis workflows and methodologies already available in R/Bioconductor packages.
 %In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).
@@
 % \section*{Description of the iSEE interface} % FM: if required, "turn on the section"?
-The layout of the iSEE user interface is built on top of the shinydashboard package. 
-The header contains three dropdown menus, related to internal iSEE diagnostics function, documentation, and additional information.
+The layout of the iSEE user interface is built using the shinydashboard package \citep{chang2018shinydashboard}.
+The header contains three dropdown menus that provide constant access to diagnostics of the current application state, documentation, and additional information, respectively.
 % TODO, FM: maybe we can/should cross-reference to the upcoming sections?
-The sidebar of iSEE contains a dropdown menu and a button for creating new plots or tables (referred to as ``panels'') in the interface.
-Panels can be added, canceled, adjusted in width and height, and re-organized, by clicking on the corresponding color-coded tabs.
-The main element in the body of iSEE is the combination of panels, generated (and optionally linked to one another) according to the users' actions.  
-Besides the package documentation and vignette, users can get familiar with the main elements of the user interface by taking an interactive tour (based on the rintrojs package []ganz[]), which will allow them to learn the basic usage mechanisms by doing.
+The sidebar of iSEE contains a menu enabling the creation of new plots and tables (referred to as ``panels'') in the interface.
+Individual color-coded tabs allow the corresponding panel to be re-ordered, adjusted in width and height, or removed from the user interface.
+The main component of the iSEE interface displays a combination of active panels controlled by the user's actions in the current session.
+Notably, individual panels can optionally be linked to one another, allowing information to be dynamically transmitted between the panels.
+In addition to the package documentation and vignette, users can familiarize themselves with the main elements of the user interface by taking an interactive tour (based on the rintrojs package []ganz[]), allowing them to learn the basic usage mechanisms by doing.
 
-\section*{Functionality}
-Currently, there are currently six different panel types that can be generated with iSEE.
+\section*{Functionalities}
+Currently, six different panel types can be generated with iSEE.
 For each panel, three different set of parameters are available in collapsible boxes.

### 2018-04-28 manuscript/main.tex
edit on the reduced dim plots

@@
 \section*{Functionalities}
-Currently, six different panel types can be generated with iSEE.
-For each panel, three different set of parameters are available in collapsible boxes.
-Data parameters control settings specific to each type of plot; Visual parameters specify parameters that will determine the aspect of the plot (coloring, point features, ...); Selection parameters relate to the point selections and link relationships to other plots.
+Currently, six different types of panels can be generated with iSEE.
+For each panel, three different set of parameters are available in collapsible boxes: namely, Data parameters control the sources of information specific to each type of plot; Visual parameters direct the appearance of the data points (e.g.
+color, opacity, size); and Selection parameters adjust the behaviour of point selections transmitted between linked panels.
+
+Individual panel types include:
 
 \begin{itemize}
-\item Reduced dimension plots -  Display any reduced dimension results (e.g., PCA, t-SNE), extracted from the reducedDim slot of a SingleCellExperiment object is supplied to iSEE, enabling efficient exploration of high-dimensional datasets.
+\item Reduced dimension plots -  Display any two dimensions from individual precomputed reduced dimension matrices (e.g., PCA, t-SNE) available in the reducedDim slot of a SingleCellExperiment object supplied to iSEE.
+Notably, the capacity of SingleCellExperiment objects to store any number of reduced dimension matrices enables efficient exploration of high-dimensional datasets.
 \item Column data plots - Visualize sample metadata stored in the SummarizedExperiment object, whereas plots can assume various forms, depending on the nature of the data on the x- and y-axes.

### 2018-04-28 manuscript/main.tex
remove unnecessary word

@@
 \begin{itemize}
-\item Reduced dimension plots -  Display any two dimensions from individual precomputed reduced dimension matrices (e.g., PCA, t-SNE) available in the reducedDim slot of a SingleCellExperiment object supplied to iSEE.
+\item Reduced dimension plots -  Display any two dimensions from precomputed reduced dimension matrices (e.g., PCA, t-SNE) available in the reducedDim slot of a SingleCellExperiment object supplied to iSEE.
 Notably, the capacity of SingleCellExperiment objects to store any number of reduced dimension matrices enables efficient exploration of high-dimensional datasets.

### 2018-04-28 manuscript/main.tex
column data plot

@@
 Notably, the capacity of SingleCellExperiment objects to store any number of reduced dimension matrices enables efficient exploration of high-dimensional datasets.
-\item Column data plots - Visualize sample metadata stored in the SummarizedExperiment object, whereas plots can assume various forms, depending on the nature of the data on the x- and y-axes.
+\item Column data plots - Visualize sample metadata stored in the colData slot of SummarizedExperiment object.
+Conveniently, both categorical and continuous covariates are seamlessly supported, with the geometry of the plot layers automatically switching between scatter plots, violin plots, or bi-categorical grid plots, according to the nature of the data on both axes.
 \item Feature expression plots - Visualize the expression values for a particular feature (e.g., a gene) across the samples on the y-axis, often resulting in (grouped) violin plots or scatter plots, if the X axis is set to categorical or continuous variables, respectively.

### 2018-04-28 manuscript/main.tex
feature expression plot

@@
 Conveniently, both categorical and continuous covariates are seamlessly supported, with the geometry of the plot layers automatically switching between scatter plots, violin plots, or bi-categorical grid plots, according to the nature of the data on both axes.
-\item Feature expression plots - Visualize the expression values for a particular feature (e.g., a gene) across the samples on the y-axis, often resulting in (grouped) violin plots or scatter plots, if the X axis is set to categorical or continuous variables, respectively.
+\item Feature expression plots - Visualize the expression values for a particular feature (e.g., gene, protein) within individual samples.
+This type of plot adopts either a violin or a scatter plot geometry, according to the nature of the second covariate.
 \item Row statistics tables - Present the values of the rowData slot for the SummarizedExperiment object.

### 2018-04-28 manuscript/main.tex
refined featExprPlot

@@
 Conveniently, both categorical and continuous covariates are seamlessly supported, with the geometry of the plot layers automatically switching between scatter plots, violin plots, or bi-categorical grid plots, according to the nature of the data on both axes.
-\item Feature expression plots - Visualize the expression values for a particular feature (e.g., gene, protein) within individual samples.
-This type of plot adopts either a violin or a scatter plot geometry, according to the nature of the second covariate.
+\item Feature expression plots - Visualize expression values from any assay stored in the SummarizedExperiment object for a particular feature (e.g., gene, protein) within individual samples.
+This type of plot often adopts a violin or scatter plot geometry, according to the nature of the second covariate.
 \item Row statistics tables - Present the values of the rowData slot for the SummarizedExperiment object.

### 2018-04-28 manuscript/main.tex
row stat table

@@
 This type of plot often adopts a violin or scatter plot geometry, according to the nature of the second covariate.
-\item Row statistics tables - Present the values of the rowData slot for the SummarizedExperiment object.
-A typical usage for such tables is to link to other plots, in order to determine the features to use for plotting, or coloring.
+\item Row statistics tables - Present the contents of the rowData slot for the SummarizedExperiment object.
+These tables are particularly helpful when dynamically linked with other panels, both as summary tables of selected genes and sources of selections to visualise or color in target plots
 \item Row data plots - Visualize the information stored in the rowData slot of a SummarizedExperiment object, in a behavior that follows the implementation of column data plots.

### 2018-04-28 manuscript/main.tex
row data plots

@@
 \item Row statistics tables - Present the contents of the rowData slot for the SummarizedExperiment object.
-These tables are particularly helpful when dynamically linked with other panels, both as summary tables of selected genes and sources of selections to visualise or color in target plots
-\item Row data plots - Visualize the information stored in the rowData slot of a SummarizedExperiment object, in a behavior that follows the implementation of column data plots.
+These tables are particularly helpful when dynamically linked with other panels, both as summary tables of selected genes and sources of selections to visualise or color in target plots.
+\item Row data plots - Visualize the information stored in the rowData slot of a SummarizedExperiment object, following an implementation similar to column data plots.
 \item Heat maps - Visualize overviews of the data at hand, in the form of color-coded expression matrices, where features such as genes are the rows, and samples are the columns.

### 2018-04-28 manuscript/main.tex
heat map

@@
 \item Row data plots - Visualize the information stored in the rowData slot of a SummarizedExperiment object, following an implementation similar to column data plots.
-\item Heat maps - Visualize overviews of the data at hand, in the form of color-coded expression matrices, where features such as genes are the rows, and samples are the columns.
-Selection of features is possible from dedicated widgets, which also allow subsetting based on other panels.
+\item Heat maps - Simultaneously visualize assay data and sample metadata, in the form of color-coded expression matrix where rows represent features and columns represent samples.
+As such, heat maps may receive both row and column data inputs from appropriately linked panels, to control the subsetting and highlighting of individual features and samples selected in the respective source panels.
 \end{itemize}

### 2018-04-28 manuscript/main.tex
data point coloring

@@
 
-Coloring of points (i.e., samples) on each plot can be achieved in different ways, where the default is no color scheme and results in black data points.
-Any column from the metadata of the SummarizedExperiment can be used, and the same is true for the expression values of a particular feature.
+The color of individual data points (i.e., samples) can be controlled in different ways in each plot.
+The default is no color scheme, with the same color (default, black) being applied uniformly to all data points.
+However, any column from the metadata available in the SummarizedExperiment may be used, as well as any any assay data.
 The plotting routines automatically adjust the scale to use, based on the chosen variable.

### 2018-04-28 manuscript/main.tex
dynamic links

@@
 A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
-Users can define arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting tables and plots highlight the corresponding data points in receiving panels.
-This feature facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to one or more feature expression plots.
-The linking paradigm extends to multiple plots whereby a plot can transmit to multiple receivers, or a receiving plot can itself transmit its own selection to another plot.
-This allows users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}, with the advantage of being extensible to any (meta)data present in a SummarizedExperiment object.
-Information can similarly be transferred between tables and plots, e.g., selecting a row determines how points are coloured in receiving plots.
+Users can define and reorganize arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting panels control the inclusion and appearance of the corresponding data points in receiving panels.
+This feature facilitates exploration of the relationships between different aspects of the data.
+For example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding -- this is achieved by transmitting information from the reduced dimension plot to any number of feature expression plots.
+
+Importantly, the linking paradigm extends to multiple panels, whereby a panel can transmit to multiple receivers, and a receiving panel can transmit its own selection to another plot.
+Chains of linked plots allow users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}.
+Nevertheless, iSEE augments this functionality with the advantage of integrating filters based on any feature and column metadata present in a SummarizedExperiment object.
+Lastly, both tables and plot panels may be used as sources and targets of dynamically linked selections (e.g., selecting a table row to control the coloring of data points in a plot).
 

### 2018-04-28 manuscript/main.tex
stub of author contribution

@@
 \subsection*{Author contributions}
-In order to give appropriate credit to each author of an article, the individual
-contributions of each author to the manuscript should be detailed in this section. We
-recommend using author initials and then stating briefly how they contributed.
+All authors contributed conceptually and technically to the development and maintenance of large sections of the package code.
+All authors wrote and approved the final manuscript.
 

### 2018-04-28 manuscript/main.tex
first go at specific author contributions
not filters

@@
 Chains of linked plots allow users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}.
-Nevertheless, iSEE augments this functionality with the advantage of integrating filters based on any feature and column metadata present in a SummarizedExperiment object.
+Nevertheless, iSEE augments this functionality with the advantage of integrating selections based on any feature and column metadata present in a SummarizedExperiment object.
 Lastly, both tables and plot panels may be used as sources and targets of dynamically linked selections (e.g., selecting a table row to control the coloring of data points in a plot).
@@
 All authors contributed conceptually and technically to the development and maintenance of large sections of the package code.
+A.T.L.L. and F.M. developed the application and data infrastructure.
+K.R.A and C.S. developed the plotting framework.
+F.M. developed the code tracking and user interface.
+All authors contributed to testing the application on all data sets.
 All authors wrote and approved the final manuscript.

### 2018-04-28 manuscript/main.tex
startup configuration

@@
 
-Additional parameters, provided when launching the iSEE application, can customize the startup state of the application
-The number and identity of the panels, as well as their inter-relationships can also be specified at initialization, thus avoiding the need to click through the choices to obtain the desired panel setup.
-An example of such customization is provided in %% TODO %%
-Users can refer to the documentation of ?defaults for the full list of values that can be specified in iSEE.
+Instances of iSEE applications may be individually pre-configured to alter the startup layout and state of individual panel inputs.
+Conveniently, this feature makes it possible to showcase salient features of any data set immediately upon startup, avoiding the need for any unnecessary user input.
+In particular, startup settings include the number of panels of each type, their position, width, and height in the interface, individual variables used for each axis and color scale, as well as dynamic links between panels and even brushes and lasso selections may all be configured and saved as an R script that may be run or shared with collaborators for this purpose.
+Example of such customization are readily demonstrated by the tours, available at
+\url{https://github.com/LTLA/iSEE2018/tree/master/tours}. %% TODO: is that redundant of the later description of tours? %%
+A comprehensive list of configurable startup values is available in the ?defaults documentation.
 

### 2018-04-28 manuscript/main.tex
ExperimentColorMap section

@@
 iSEE also offers a host of other customizable features to improve data exploration.
-Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each metadata variable and assay type across all panels in the application.
-% TODO: KRA develop here
+To complement the default discrete and continuous color scales borrowed from the ggplot2 \citep{wickham2009ggplot2} and viridis packages \citep{garnier2018viridis}, respectively, iSEE implements the ExperimentColorMap class, that acts as a container capable of accomodating any number of arbitrary color maps.
+Color maps are stored as functions that require a single argument, namely the number of colors that they should return.
+A notable exception are discrete color maps, that may choose to ignore their argument in order to consistently return a constant vector of named colors corresponding to the levels of a known factor.
+Color maps stored in ExperimentColorMap objects can be named and retrieved through intuitive accessor functions that follow a naming convention mirroring that of SummarizedExperiment objects (i.e., assayColorMap, colDataColorMap, rowDataColorMap), and thereby facilitates integration within Bioconductor workflows.
+
+Notably, iSEE uses a ExperimentColorMap object to coordinate the color scale applied to every assay, column data, and row data variable across all plots of a session.
+The default ExperimentColorMap does not contain any custom color map, and leverages an elegant fallback mechanism to consistently return a discrete or continuous colormap, even for undefined covariates.
+
+Color maps may be overriden at three different levels:
+
+\begin{itemize}
+  \item Each individual assay, column data, and row data covariate can be assigned its own distinct color map. Those color maps in the assays, colData, and rowData slots, respectively.
+  \item Shared colormaps can be defined for all assays, all column data, and all row data covariates, respectively.
+These shared color maps come into effect for variables that do not have an individual color map.
+  \item Globally shared color maps can be defined for all discrete or continuous covariate that do not have color maps defined at the first two levels.
+\end{itemize}
 

### 2018-04-28 manuscript/main.tex
minor edit
tweaks to the code tracking section

@@
 
-Instances of iSEE applications may be individually pre-configured to alter the startup layout and state of individual panel inputs.
+Instances of iSEE applications may be individually pre-configured to alter the startup layout and state of individual panels.
 Conveniently, this feature makes it possible to showcase salient features of any data set immediately upon startup, avoiding the need for any unnecessary user input.
@@
 iSEE also automatically memorizes the exact R code that was used to generate every plot and the existing links, extending previous work by \cite{marini2016interrepro}.
-A modal popup window, with a shinyAce-based text editor, where the code is formatted and displayed with syntax highlighting, is opened when clicking on the `Extract the R code'' button from the diagnostics dropdown menu in the header.
-This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.
-Additionally, the ``Examine panel chart'' functionality, in the iSEE diagnostics dropdown menu, returns a graph representation of the existing links and point selection among the open panels.
-This can be very useful in sessions that include a large number of panels.
-Clicking on the ``Display panel settings'' button from the same dropdown menu returns the required lines of code that underlie the current setup of currently open panels.
+The full code is accessible to users at any time, using the `Extract the R code'' button from the diagnostics dropdown menu in the header.
+A modal window displays the formatted code in a shinyAce-based text editor with syntax-highlighting and copy-pasting capabilities, allowing users to easily reproduce the results of any exploratory analysis, simply by integrating the code reported by iSEE into their own scripts.
+
+Additionally, the ``Examine panel chart'' functionality, in the iSEE diagnostics dropdown menu, displays a graphical representation of the existing linked point selections between the visible panels.
+This can be very useful in sessions that include a large number of panels, and complex 'gating' strategies.
+Lastly, the ``Display panel settings'' button from the same dropdown menu returns the required lines of code that underlie the current setup of currently open panels, that may be used in startup scripts to launch an iSEE instance in any prefered layout.
 

### 2018-04-28 manuscript/main.tex
app tours and development

@@
 \subsection*{App tours and deployment}
-After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
-This normally requires the setup of a Shiny Server, % TODO: FM, do we want to provide some basic instructions on how to do it?,
-as well as the preparation of the dataset at hand in form of a SummarizedExperiment - which many workflows available in Bioconductor adopt as main representation.
-Tours can be provided to the iSEE function in the form of simple data frames, containing the list of DOM elements in the Shiny app to anchor the corresponding descriptions.
+
+After completion of an analysis, users may deploy instances of iSEE applications on web-servers, and include a bespoke step-by-step ``tour'' of their data set, guiding the audience through an examination of the salient features in the data.
+
+This normally requires the setup of a Shiny Server, % TODO: FM, do we want to provide some basic instructions on how to do it?, % KRA: I don't think so, at least not in the paper.
+However, documentation and skill will be a limiting factor to encourage users
+as well as the preparation of the data set at hand in the form of a SummarizedExperiment - which many workflows available in Bioconductor adopt as main representation.
+To include results obtained from any number of dimensionality reduction algorithms, the object must be further coerced to the SingleCellExperiment class, that makes the appropriate slot available.
+
+Tours can be provided to the iSEE function in the form of simple data frames containing a list of Document Object Model (DOM) elements in the Shiny app used to anchor the corresponding associated message ultimately displayed to the user.
 Examples of such tours can be found in the inst/extdata folder of the iSEE package, as well as in the GitHub repository of this manuscript, \url{https://github.com/LTLA/iSEE2018/tree/master/tours}.

### 2018-04-28 manuscript/main.tex
demo and tours

@@
 \section*{Use cases} % FM: used often in other software articles in F1000res
-To demonstrate the functionality and flexibility of the iSEE package, we applied our method to a number of publicly available datasets.
+To demonstrate the functionalities and flexibility of the iSEE package, we applied our method to a number of publicly available datasets.
 Details are provided in the tour subfolder of the GitHub repository for this manuscript (\url{https://github.com/LTLA/iSEE2018}), and for each dataset a combination of \_app.R and \_tour.txt is provided.
-An instance of iSEE running on the respective datasets is provided as a stand-alone webpages, for demonstration purposes without the need of software installation.
+An instance of iSEE running on the respective datasets is provided as a stand-alone webpages, for demonstration purposes without the need for any software installation.
 

### 2018-04-28 manuscript/main.tex
R code. Layout web link to allen app

@@
 \usepackage{url}
+\usepackage{listings}
 
@@
 We demonstrate the application of iSEE to examine a subset of 379 cells from of mouse visual cortex \citep{tasic2016allen}.
-This example can be reproduced also by calling
 
+Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE/}
+
+This particular example can be reproduced locally after installation of the iSEE package with the following code:
+
+\begin{lstlisting}[language=R]
 library(iSEE)
 example(iSEE, ask = F)
-
-% TODO: embed the allen demo data set
-\url{http://shiny.imbei.uni-mainz.de:3838/iSEE/}
+\end{lstlisting}
 

### 2018-04-28 manuscript/main.tex
better layout of link
removed superfluous word
RNA-seq

@@
 We demonstrate the application of iSEE to examine a subset of 379 cells from of mouse visual cortex \citep{tasic2016allen}.
-
-Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE/}
-
-This particular example can be reproduced locally after installation of the iSEE package with the following code:
+This example can be reproduced locally after installation of the iSEE package with the following code:
 
@@
 
+Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE/}
+
 \subsection*{Single-cell droplet-based RNA-seq}
 % TODO: better title, or do we drop the allen section altogether (above)?
-We also demonstrate the use of iSEE to explore a larger data set of 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively}.
+We also demonstrate the use of iSEE to explore a larger data set of 4,000 PBMC single-cell RNA-sequencing dataset from 10X Genomics \citep{zheng2017massively}.
 % TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?

### 2018-04-28 manuscript/main.tex
From Kevin: More URL links and conclusion. Added F1000 style.

@@
 \title{iSEE: Interactive SummarizedExperiment Explorer}
-\titlenote{The title should be detailed enough for someone to know whether the article would be of interest to them, but also concise.
-Please ensure the broadness and claims within the title are appropriate to the content of the article itself.}
 \author[1,$\dagger$]{Kevin Rue-Albrecht}
@@
 \begin{abstract}
-
-\textbf{Summary:} Data exploration is critical to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
+Data exploration is critical to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
 However, most existing tools for interactive visualization are limited to specific assays or analyses.
@@
 and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
-We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.\\
-\textbf{Availability:} iSEE is publicly available as an R package from the open-source Bioconductor project (\url{https://bioconductor.org/packages/iSEE}).\\
-\textbf{Contact:} aaron.lun@cruk.cam.ac.uk\\
-\textbf{Supplementary information:} Supplementary data are available online.
+We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.
 
@@
 Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
-To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions, microarray probe sets).
+To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions, microarray probe sets).\\
 
@@
 % TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
-% embed the allen demo data set
-\url{http://shiny.imbei.uni-mainz.de:3838/iSEE_PBMC4k/}
+
+% TODO: embed the allen demo data set
+Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_PBMC4k/}
 
@@
 % TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
+
 % TODO: embed the TCGA demo data set
-\url{http://shiny.imbei.uni-mainz.de:3838/iSEE_TCGA/}
+Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_TCGA/}
 
 \subsection*{CyTOF}
-Lastly, we also showcase the use of iSEE to vizualise and refine gating a gating analysis, followed by the investigation of a functional marker between two sample groups, using a published mass cytometry (CyTOF) data set \citep{bodenmiller2012cytof}.
+Lastly, we also showcase the use of iSEE to vizualise and refine a gating analysis, followed by the investigation of a functional marker between two sample groups, using a published mass cytometry (CyTOF) data set \citep{bodenmiller2012cytof}.
 % TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
 
-\url{http://shiny.imbei.uni-mainz.de:3838/iSEE_cytof/}
+Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_cytof/}
 
@@
 % TODO: sell the idea that for every dataset it is possible to explore and showcase feats of interest. Something like...
-We expect the functionality and flexibility of the iSEE package to be essential when exploring and showcasing relevant features of any dataset which can be represented in the form of a SummarizedExperiment object, in such a way that a public instance of iSEE can ideally accompany any publication, e.g.
-reproducing the manuscript figures, but also allowing any user to foster hypothesis generation with efficient interactive and reproducible data analysis.
+We expect the functionalities and flexibility of the iSEE package to be essential for exploring and showcasing relevant features of any data set that can be represented in the form of a SummarizedExperiment object.
+In particular, we expect public instances of iSEE to ideally accompany publications, as they enable authors to reproduce published figures and showcase salient aspects of their work using guided tours, while fostering hypothesis generation through transparent, interactive, reproducible, and efficient data analysis by the broader reader audience.
 

### 2018-04-28 manuscript/main.tex
Reorganized functionalities section to remove redundancies, detail.

@@
 \section*{Introduction}
-Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics.
+Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as genomics.
 Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
-To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions, microarray probe sets).\\
+To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).\\
 
@@
 Moreover, standalone tools such as the Loupe Cell Browser from 10X Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}.
-This complicates any coordinated use of these tools with a reproducible, transparent, and statistically rigorous analysis.
+This complicates any coordinated use of these tools with a reproducible, transparent, and statistically rigorous analysis.\\
+
+Here, we present the iSEE software package for interactive data exploration.
+iSEE is implemented in R using the Shiny framework \citep{chang2017shiny} and exploits data structures from the open-source Bioconductor project \citep{gentleman2004bioconductor}, specifically the SummarizedExperiment class.
+iSEE allows users to simultaneously visualize multiple aspects of a given dataset, including experimental data, metadata and analysis results.
+Dynamic linking and point selection facilitate the flexible exploration of interactions between different data aspects.
+Additional functionalities include code tracking, intelligent downsampling of large datasets, custom colour scale specification and tour construction.
+We demonstrate the capabilities of iSEE by applying it to a diverse range of real datasets.
+
+\section*{Taking a SummarizedExperiment input}
+The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
+Each instance of the SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where rows and columns represent genomic features and biological samples, respectively.
+For instance, individual assays may represent gene expression matrices of raw counts along with different normalization strategies.
+In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; those may include experimental metadata as well as any analysis results.\\
+
+The flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem.
+SummarizedExperiment objects are currently used in analysis pipelines for RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C data \citep{lun2016infrastructure}, amongst others.
+Package developers can also easily use the base SummarizedExperiment class to derive new bespoke classes for particular applications, such as the SingleCellExperiment class for single-cell 'omics data.
+By accepting SummarizedExperiment objects as input, iSEE immediately offers interactive visualization for a variety of data modalities.
+This complements the state-of-the-art analysis workflows and methodologies already available in R/Bioconductor packages. \\
+
+\section*{Interface implementation}
+
+\subsection*{Using a multi-panel layout}
+All data aspects stored in a SummarizedExperiment can be readily examined alongside one another in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
+The interface layout is built using the shinydashboard package \citep{chang2018shinydashboard}, using colour-coded panels to visualize each data aspect.
+Individual panel types include:
+
+\begin{itemize}
+\item Column data plots, for visualizing sample metadata stored in the colData slot of SummarizedExperiment object. 
+\item Feature expression plots, for visualizing expression values for a particular feature (e.g., gene) across samples from any assay in the SummarizedExperiment object.
+Despite its name, this plot type is applicable to data modalities beyond gene expression.
+\item Row statistics tables, to present the contents of the rowData slot for the SummarizedExperiment object.
+\item Row data plots, for visualizing sample metadata stored in the rowData slot of a SummarizedExperiment object.
+\item Heat maps, to visualize assay data for multiple features where samples are ordered by one or more colData fields.
+\item Reduced dimension plots, which display any two dimensions from pre-computed dimensionality reduction results (e.g., from PCA or $t$-SNE).
+These results are taken from the reducedDim slot of any SingleCellExperiment object supplied to iSEE. 
+\end{itemize}
 
@@
 \caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data.
+Panels of each type are colour-coded for ease of use.
 New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
 Panel types are available to visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
-For these panel types, each sample is represented as a point, which can be coloured according to sample-level metadata or feature-level observations.
-Depending on whether the variables of interest are categorical or numeric, panels will automatically switch between scatter plots, violin plots or ``rectangle plots'' (where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of combination).
 Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
@@
 
-\section*{Implementation}
-Here, we present the iSEE software package for interactive data exploration.
-iSEE is implemented in R using the Shiny framework \citep{chang2017shiny} and exploits data architectures from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
-The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
-The SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where rows and columns represent genomic features and biological samples, respectively.
-For instance, individual assays may represent gene expression matrices of raw counts along with different normalization strategies.
-In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; those may include experimental metadata as well as any analysis results.
-All of these data aspects can be readily examined alongside one another in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
+Each sample is represented as a point in column data, feature expression and reduced dimension plots.
+Similarly, each feature is represented by a point in row data plots.
+For these panel types, a scatter plot is automatically produced if the selected variables on the x- and y-axes are both continuous.
+If exactly one variable is categorical, points are grouped by the categorical levels and a (horizontal) violin plot is produced with points scattered within each violin.
+If both variables are categorical, a ``rectangle plot'' is produced where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of that combination.
+Points are scattered randomly within each rectangle.
+For ease of interpretation, the rectangle plot collapses to a mirrored bar plot when one of the categorical variables only has one level.
 
-Notably, the flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
-As a consequence, the widespread adoption of the SummarizedExperiment class and its specialized extensions such as the SingleCellExperiment class enables iSEE to offer immediate interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis workflows and methodologies already available in R/Bioconductor packages.
-%In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).
-% KRA: any other child class like SingleCellExperiment worth mentioning? e.g. the BSseq class
+\subsection*{Custom panel colouring}
+Sample-based points can be coloured according to the values of any sample-level metadata field in the colData or by the assay values of a selected feature.
+Similarly, feature-based points can be coloured according to any feature-level metadata field in the rowData.
+Heatmaps are coloured according to the expression values of the selected features in the chosen assay,
+with additional colouring of each of the colData fields used to order the samples.
+In all cases, the variable to use for colouring can be dynamically selected for each plot.
+This enables users to easily examine relationships between different variables in a single plot. \\
 
-% \section*{Description of the iSEE interface} % FM: if required, "turn on the section"?
-The layout of the iSEE user interface is built using the shinydashboard package \citep{chang2018shinydashboard}.
-The header contains three dropdown menus that provide constant access to diagnostics of the current application state, documentation, and additional information, respectively.
-% TODO, FM: maybe we can/should cross-reference to the upcoming sections?
-The sidebar of iSEE contains a menu enabling the creation of new plots and tables (referred to as ``panels'') in the interface.
-Individual color-coded tabs allow the corresponding panel to be re-ordered, adjusted in width and height, or removed from the user interface.
-The main component of the iSEE interface displays a combination of active panels controlled by the user's actions in the current session.
-Notably, individual panels can optionally be linked to one another, allowing information to be dynamically transmitted between the panels.
-In addition to the package documentation and vignette, users can familiarize themselves with the main elements of the user interface by taking an interactive tour (based on the rintrojs package []ganz[]), allowing them to learn the basic usage mechanisms by doing.
-
-\section*{Functionalities}
-Currently, six different types of panels can be generated with iSEE.
-For each panel, three different set of parameters are available in collapsible boxes: namely, Data parameters control the sources of information specific to each type of plot; Visual parameters direct the appearance of the data points (e.g.
-color, opacity, size); and Selection parameters adjust the behaviour of point selections transmitted between linked panels.
-
-Individual panel types include:
-
-\begin{itemize}
-\item Reduced dimension plots -  Display any two dimensions from precomputed reduced dimension matrices (e.g., PCA, t-SNE) available in the reducedDim slot of a SingleCellExperiment object supplied to iSEE.
-Notably, the capacity of SingleCellExperiment objects to store any number of reduced dimension matrices enables efficient exploration of high-dimensional datasets.
-\item Column data plots - Visualize sample metadata stored in the colData slot of SummarizedExperiment object.
-Conveniently, both categorical and continuous covariates are seamlessly supported, with the geometry of the plot layers automatically switching between scatter plots, violin plots, or bi-categorical grid plots, according to the nature of the data on both axes.
-\item Feature expression plots - Visualize expression values from any assay stored in the SummarizedExperiment object for a particular feature (e.g., gene, protein) within individual samples.
-This type of plot often adopts a violin or scatter plot geometry, according to the nature of the second covariate.
-\item Row statistics tables - Present the contents of the rowData slot for the SummarizedExperiment object.
-These tables are particularly helpful when dynamically linked with other panels, both as summary tables of selected genes and sources of selections to visualise or color in target plots.
-\item Row data plots - Visualize the information stored in the rowData slot of a SummarizedExperiment object, following an implementation similar to column data plots.
-\item Heat maps - Simultaneously visualize assay data and sample metadata, in the form of color-coded expression matrix where rows represent features and columns represent samples.
-As such, heat maps may receive both row and column data inputs from appropriately linked panels, to control the subsetting and highlighting of individual features and samples selected in the respective source panels.
-\end{itemize}
-
-The color of individual data points (i.e., samples) can be controlled in different ways in each plot.
-The default is no color scheme, with the same color (default, black) being applied uniformly to all data points.
-However, any column from the metadata available in the SummarizedExperiment may be used, as well as any any assay data.
-The plotting routines automatically adjust the scale to use, based on the chosen variable.
+By default, colour maps for categorical and continuous variables are taken from the ggplot2 \citep{wickham2009ggplot2} and viridis packages \citep{garnier2018viridis}, respectively.
+However, iSEE also implements the ExperimentColorMap class, which allows users to specify arbitrary colour maps for particular variables.
+Each colour map is a function that returns a vector of distinct colours of a specified length, and will be called whenever the associated variable is used for point colouring in a particular panel.
+The returned colours will be mapped to factor levels for categorical variables, or used in colour interpolation for continuous variables.
+For categorical variables, the function may also return a constant vector of named colors corresponding to the levels of a known factor.
+Colour maps can be specified for individual variables; for all assays, all column data variables, or all row data variables (with different functions for continuous or categorical variables); or for all categorical or continuous variables.
+This provides a convenient yet flexible mechanism for customization of colouring schemes within the interface.
 
@@
 This feature facilitates exploration of the relationships between different aspects of the data.
-For example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding -- this is achieved by transmitting information from the reduced dimension plot to any number of feature expression plots.
+For example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding -- this is achieved by selecting points in a reduced dimension plot (using the standard rectangular brush or a lasso) and transmitting that selection to any number of feature expression plots.\\
 
-Importantly, the linking paradigm extends to multiple panels, whereby a panel can transmit to multiple receivers, and a receiving panel can transmit its own selection to another plot.
+This linking paradigm extends to multiple panels, whereby a panel can transmit to multiple receivers, and a receiving panel can transmit its own selection to another plot.
 Chains of linked plots allow users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}.
-Nevertheless, iSEE augments this functionality with the advantage of integrating selections based on any feature and column metadata present in a SummarizedExperiment object.
-Lastly, both tables and plot panels may be used as sources and targets of dynamically linked selections (e.g., selecting a table row to control the coloring of data points in a plot).
+With iSEE, this concept is extended to any assay data, feature-level or sample-level metadata present in a SummarizedExperiment object, providing a powerful framework for interrogating multiple interactions between data aspects.
+Row statistics tables can also transmit to various plot types, by selecting a table row to control the coloring of sample-based points;
+or by defining a subset of features to visualize in a heatmap.
+Furthermore, row data plots can transmit to row statistics tables, whereby selection of points in the former will subset the latter.
 
-\subsection*{Advanced features and customization}
+\subsection*{Code tracking and reproducibility}
 
-Instances of iSEE applications may be individually pre-configured to alter the startup layout and state of individual panels.
-Conveniently, this feature makes it possible to showcase salient features of any data set immediately upon startup, avoiding the need for any unnecessary user input.
-In particular, startup settings include the number of panels of each type, their position, width, and height in the interface, individual variables used for each axis and color scale, as well as dynamic links between panels and even brushes and lasso selections may all be configured and saved as an R script that may be run or shared with collaborators for this purpose.
-Example of such customization are readily demonstrated by the tours, available at
-\url{https://github.com/LTLA/iSEE2018/tree/master/tours}. %% TODO: is that redundant of the later description of tours? %%
-A comprehensive list of configurable startup values is available in the ?defaults documentation.
+iSEE automatically memorizes the exact R code that was used to generate every plot, extending previous work by \cite{marini2016interrepro}.
+This code is fully accessible to users at any time during the run-time of the interface.
+By integrating the code reported by iSEE into their own scripts, users to easily reproduce the results of any exploratory analysis.
+Similarly, the code required to reproduce the current state of the interface can also be generated.
+This can be used in startup scripts to launch an iSEE instance in any preferred layout, including the panel organization, variable selection, colouring schemes, links between panels and even individual brushes and lasso selections.
 
-\subsection*{ExperimentColorMap: custom color maps}
-
-iSEE also offers a host of other customizable features to improve data exploration.
-To complement the default discrete and continuous color scales borrowed from the ggplot2 \citep{wickham2009ggplot2} and viridis packages \citep{garnier2018viridis}, respectively, iSEE implements the ExperimentColorMap class, that acts as a container capable of accomodating any number of arbitrary color maps.
-Color maps are stored as functions that require a single argument, namely the number of colors that they should return.
-A notable exception are discrete color maps, that may choose to ignore their argument in order to consistently return a constant vector of named colors corresponding to the levels of a known factor.
-Color maps stored in ExperimentColorMap objects can be named and retrieved through intuitive accessor functions that follow a naming convention mirroring that of SummarizedExperiment objects (i.e., assayColorMap, colDataColorMap, rowDataColorMap), and thereby facilitates integration within Bioconductor workflows.
-
-Notably, iSEE uses a ExperimentColorMap object to coordinate the color scale applied to every assay, column data, and row data variable across all plots of a session.
-The default ExperimentColorMap does not contain any custom color map, and leverages an elegant fallback mechanism to consistently return a discrete or continuous colormap, even for undefined covariates.
-
-Color maps may be overriden at three different levels:
-
-\begin{itemize}
-  \item Each individual assay, column data, and row data covariate can be assigned its own distinct color map. Those color maps in the assays, colData, and rowData slots, respectively.
-  \item Shared colormaps can be defined for all assays, all column data, and all row data covariates, respectively.
-These shared color maps come into effect for variables that do not have an individual color map.
-  \item Globally shared color maps can be defined for all discrete or continuous covariate that do not have color maps defined at the first two levels.
-\end{itemize}
-
-\subsection*{Linked feature annotations}
-
+\subsection*{Additional functionalities}
 Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez.
@@
 at least ENSEMBL has updated citable papers
+For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
+Users can also include a bespoke step-by-step ``tour'' of their data set via the rintrojs package \cite{ganz}, guiding the audience through an examination of the salient features in the data.
 
-\subsection*{Density-dependent downsampling for speed}
-
-For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
-
-\subsection*{Code tracking and reproducibility}
-
-iSEE also automatically memorizes the exact R code that was used to generate every plot and the existing links, extending previous work by \cite{marini2016interrepro}.
-The full code is accessible to users at any time, using the `Extract the R code'' button from the diagnostics dropdown menu in the header.
-A modal window displays the formatted code in a shinyAce-based text editor with syntax-highlighting and copy-pasting capabilities, allowing users to easily reproduce the results of any exploratory analysis, simply by integrating the code reported by iSEE into their own scripts.
-
-Additionally, the ``Examine panel chart'' functionality, in the iSEE diagnostics dropdown menu, displays a graphical representation of the existing linked point selections between the visible panels.
-This can be very useful in sessions that include a large number of panels, and complex 'gating' strategies.
-Lastly, the ``Display panel settings'' button from the same dropdown menu returns the required lines of code that underlie the current setup of currently open panels, that may be used in startup scripts to launch an iSEE instance in any prefered layout.
-
-\subsection*{App tours and deployment}
-
-After completion of an analysis, users may deploy instances of iSEE applications on web-servers, and include a bespoke step-by-step ``tour'' of their data set, guiding the audience through an examination of the salient features in the data.
-
-This normally requires the setup of a Shiny Server, % TODO: FM, do we want to provide some basic instructions on how to do it?, % KRA: I don't think so, at least not in the paper.
-However, documentation and skill will be a limiting factor to encourage users
-as well as the preparation of the data set at hand in the form of a SummarizedExperiment - which many workflows available in Bioconductor adopt as main representation.
-To include results obtained from any number of dimensionality reduction algorithms, the object must be further coerced to the SingleCellExperiment class, that makes the appropriate slot available.
-
-Tours can be provided to the iSEE function in the form of simple data frames containing a list of Document Object Model (DOM) elements in the Shiny app used to anchor the corresponding associated message ultimately displayed to the user.
-Examples of such tours can be found in the inst/extdata folder of the iSEE package, as well as in the GitHub repository of this manuscript, \url{https://github.com/LTLA/iSEE2018/tree/master/tours}.
-
-%\section*{Demonstration data sets and tours}
 \section*{Use cases} % FM: used often in other software articles in F1000res

### 2018-04-28 manuscript/main.tex
Added Ganz citation, reworded caption.

@@
 \caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data.
-Panels of each type are colour-coded for ease of use.
 New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
@@
 Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
+Panels of each type are colour-coded for ease of interpretation.
 (B) Information can be transmitted between panels according to a user-specified scheme.
@@
 For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
-Users can also include a bespoke step-by-step ``tour'' of their data set via the rintrojs package \cite{ganz}, guiding the audience through an examination of the salient features in the data.
+Users can also include a bespoke step-by-step ``tour'' of their data set via the rintrojs package \cite{ganz2016rintrojs}, guiding the audience through an examination of the salient features in the data.
 

### 2018-04-28 manuscript/main.tex
data set (Google says so) and missing 'as'

@@
 \begin{abstract}
-Data exploration is critical to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
+Data exploration is critical to the comprehension of large biological data sets generated by high-throughput assays such as sequencing.
 However, most existing tools for interactive visualization are limited to specific assays or analyses.
@@
 and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
-We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.
+We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics data sets.
 
@@
 iSEE is implemented in R using the Shiny framework \citep{chang2017shiny} and exploits data structures from the open-source Bioconductor project \citep{gentleman2004bioconductor}, specifically the SummarizedExperiment class.
-iSEE allows users to simultaneously visualize multiple aspects of a given dataset, including experimental data, metadata and analysis results.
+iSEE allows users to simultaneously visualize multiple aspects of a given data set, including experimental data, metadata and analysis results.
 Dynamic linking and point selection facilitate the flexible exploration of interactions between different data aspects.
-Additional functionalities include code tracking, intelligent downsampling of large datasets, custom colour scale specification and tour construction.
-We demonstrate the capabilities of iSEE by applying it to a diverse range of real datasets.
+Additional functionalities include code tracking, intelligent downsampling of large data sets, custom colour scale specification and tour construction.
+We demonstrate the capabilities of iSEE by applying it to a diverse range of real data sets.
 
@@
 at least ENSEMBL has updated citable papers
-For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
+For large data sets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
 Users can also include a bespoke step-by-step ``tour'' of their data set via the rintrojs package \cite{ganz2016rintrojs}, guiding the audience through an examination of the salient features in the data.
@@
 \section*{Use cases} % FM: used often in other software articles in F1000res
-To demonstrate the functionalities and flexibility of the iSEE package, we applied our method to a number of publicly available datasets.
-Details are provided in the tour subfolder of the GitHub repository for this manuscript (\url{https://github.com/LTLA/iSEE2018}), and for each dataset a combination of \_app.R and \_tour.txt is provided.
-An instance of iSEE running on the respective datasets is provided as a stand-alone webpages, for demonstration purposes without the need for any software installation.
+To demonstrate the functionalities and flexibility of the iSEE package, we applied our method to a number of publicly available data sets.
+Details are provided in the tour subfolder of the GitHub repository for this manuscript (\url{https://github.com/LTLA/iSEE2018}), and for each data set a combination of \_app.R and \_tour.txt is provided.
+An instance of iSEE running on the respective data sets is provided as a stand-alone webpages, for demonstration purposes without the need for any software installation.
 
@@
 % TODO: better title, or do we drop the allen section altogether (above)?
-We also demonstrate the use of iSEE to explore a larger data set of 4,000 PBMC single-cell RNA-sequencing dataset from 10X Genomics \citep{zheng2017massively}.
+We also demonstrate the use of iSEE to explore a larger data set of 4,000 PBMC single-cell RNA-sequencing data set from 10X Genomics \citep{zheng2017massively}.
 % TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
@@
 \subsection*{Bulk RNA-seq from TCGA}
-We also demonstrate how iSEE seamlessly scales up with larger data sets as the TCGA RNA-seq collection of measurements for 23,368 genes across 7,706 tumor samples \citep{piccolo2015TCGA}.
+We also demonstrate how iSEE seamlessly scales up with larger data sets such as the TCGA RNA-seq collection of measurements for 23,368 genes across 7,706 tumor samples \citep{piccolo2015TCGA}.
 % TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
@@
 \section*{Conclusion}
-iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
-We provide a number of demonstrations at \url{https://github.com/LTLA/iSEE2018}, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells \citep{bodenmiller2012cytof}.
+iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics data sets.
+We provide a number of demonstrations at \url{https://github.com/LTLA/iSEE2018}, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing data set from 10X Genomics \citep{zheng2017massively} and a published mass cytometry data set consisting of over 170,000 cells \citep{bodenmiller2012cytof}.
 % TODO: repurpose the conclusion if we showcase the examples in earlier sections
-% TODO: sell the idea that for every dataset it is possible to explore and showcase feats of interest. Something like...
+% TODO: sell the idea that for every data set it is possible to explore and showcase feats of interest. Something like...
 We expect the functionalities and flexibility of the iSEE package to be essential for exploring and showcasing relevant features of any data set that can be represented in the form of a SummarizedExperiment object.

### 2018-04-28 manuscript/main.tex
web page

@@
 Details are provided in the tour subfolder of the GitHub repository for this manuscript (\url{https://github.com/LTLA/iSEE2018}), and for each data set a combination of \_app.R and \_tour.txt is provided.
-An instance of iSEE running on the respective data sets is provided as a stand-alone webpages, for demonstration purposes without the need for any software installation.
+An instance of iSEE running on the respective data sets is provided as a stand-alone web pages, for demonstration purposes without the need for any software installation.
 

### 2018-04-29 manuscript/main.tex
ECM credit
Cleaned up use case descriptions.

--- 2018-04-28 manuscript/main.tex
@@
 This code is fully accessible to users at any time during the run-time of the interface.
-By integrating the code reported by iSEE into their own scripts, users to easily reproduce the results of any exploratory analysis.
-Similarly, the code required to reproduce the current state of the interface can also be generated.
+By integrating the code reported by iSEE into their own scripts, users can easily reproduce the results of any exploratory analysis.
+Similarly, the code required to reproduce the current state of the interface can also be reported.
 This can be used in startup scripts to launch an iSEE instance in any preferred layout, including the panel organization, variable selection, colouring schemes, links between panels and even individual brushes and lasso selections.
@@
 \section*{Use cases} % FM: used often in other software articles in F1000res
-To demonstrate the functionalities and flexibility of the iSEE package, we applied our method to a number of publicly available datasets.
-Details are provided in the tour subfolder of the GitHub repository for this manuscript (\url{https://github.com/LTLA/iSEE2018}), and for each dataset a combination of \_app.R and \_tour.txt is provided.
-An instance of iSEE running on the respective datasets is provided as a stand-alone webpages, for demonstration purposes without the need for any software installation.
 
-\subsection*{Single-cell plate-based RNA-seq}
-% TODO: better title, or do we drop this allen section altogether?
-We demonstrate the application of iSEE to examine a subset of 379 cells from of mouse visual cortex \citep{tasic2016allen}.
-This example can be reproduced locally after installation of the iSEE package with the following code:
-
-\begin{lstlisting}[language=R]
-library(iSEE)
-example(iSEE, ask = F)
-\end{lstlisting}
+\subsection*{Plate-based single-cell RNA sequencing}
+To demonstrate iSEE's functionality, we used it to explore a plate-based single-cell RNA sequencing (scRNA-seq) data set.
+We used a subset of data from a study involving cells from the mouse visual cortex \citep{tasic2016allen}.
+This demonstration guides the user through the main features of the iSEE interface, including the multi-panel layout, colouring options and dynamic linking.\\
 
@@
 
-\subsection*{Single-cell droplet-based RNA-seq}
-% TODO: better title, or do we drop the allen section altogether (above)?
-We also demonstrate the use of iSEE to explore a larger data set of 4,000 PBMC single-cell RNA-sequencing dataset from 10X Genomics \citep{zheng2017massively}.
-% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
+\subsection*{Droplet-based single-cell RNA sequencing}
+We applied iSEE to a larger scRNA-seq data set involving 4,000 peripheral blood mononuclear cells (PBMCs), generated by 10X Genomics \citep{zheng2017massively}.
+This demonstration explores the differences between different methods for distinguishing cells from empty droplets in droplet-based scRNA-seq protocols \citep{lun2018distinguishing}.\\
 
-% TODO: embed the allen demo data set
 Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_PBMC4k/}
 
-\subsection*{Bulk RNA-seq from TCGA}
-We also demonstrate how iSEE seamlessly scales up with larger data sets as the TCGA RNA-seq collection of measurements for 23,368 genes across 7,706 tumor samples \citep{piccolo2015TCGA}.
-% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
+\subsection*{Bulk RNA sequencing from TCGA}
+We applied iSEE to bulk RNA sequencing data from The Cancer Genome Atlas (TCGA) project, using a subset of expression profiles involving 7,706 tumor samples \citep{piccolo2015TCGA}.
+This demonstration examines the elevation of \textit{HER2} expression in breast cancer samples.\\
 
-% TODO: embed the TCGA demo data set
 Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_TCGA/}
 
-\subsection*{CyTOF}
-Lastly, we also showcase the use of iSEE to vizualise and refine a gating analysis, followed by the investigation of a functional marker between two sample groups, using a published mass cytometry (CyTOF) data set \citep{bodenmiller2012cytof}.
-% TODO: take out the juice from what the tour does, and maybe one figure/screenshot in each case?
+\subsection*{Mass cytometry}
+Finally, we explored a mass cytometry study involving PBMCs from multiple donors before and after stimulation with BCR/FcR-XL \citep{bodenmiller2012cytof}.
+We used iSEE to vizualise and refine a gating analysis to obtain B cells, and to investigate differences in expression of the functional marker pS6 after stimulation.\\
 

### 2018-04-29 manuscript/main.tex

@@
 F.M. developed the code tracking and user interface.
+K.R.A designed and developed the ExperimentColorMap class.
 All authors contributed to testing the application on all data sets.

### 2018-04-29 manuscript/main.tex
Added numbers to all use cases.

@@
 \subsection*{Plate-based single-cell RNA sequencing}
-To demonstrate iSEE's functionality, we used it to explore a plate-based single-cell RNA sequencing (scRNA-seq) data set.
-We used a subset of data from a study involving cells from the mouse visual cortex \citep{tasic2016allen}.
-This demonstration guides the user through the main features of the iSEE interface, including the multi-panel layout, colouring options and dynamic linking.\\
+To demonstrate iSEE's functionality, we used it to explore a plate-based single-cell RNA sequencing (scRNA-seq) data set involving 379 cells from the mouse visual cortex \citep{tasic2016allen}.
+This demonstration guides the user through the main features of the iSEE interface including the multi-panel layout, colouring and dynamic linking.\\
 
@@
 \subsection*{Mass cytometry}
-Finally, we explored a mass cytometry study involving PBMCs from multiple donors before and after stimulation with BCR/FcR-XL \citep{bodenmiller2012cytof}.
+Finally, we explored a mass cytometry study involving 150,000 PBMCs from multiple donors before and after stimulation with BCR/FcR-XL \citep{bodenmiller2012cytof}.
 We used iSEE to vizualise and refine a gating analysis to obtain B cells, and to investigate differences in expression of the functional marker pS6 after stimulation.\\

### 2018-04-29 manuscript/main.tex
Cleaned up conclusion and rearranged sundries.

@@
 \section*{Conclusion}
-iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics data sets.
-We provide a number of demonstrations at \url{https://github.com/LTLA/iSEE2018}, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing data set from 10X Genomics \citep{zheng2017massively} and a published mass cytometry data set consisting of over 170,000 cells \citep{bodenmiller2012cytof}.
-% TODO: repurpose the conclusion if we showcase the examples in earlier sections
-% TODO: sell the idea that for every data set it is possible to explore and showcase feats of interest. Something like...
-We expect the functionalities and flexibility of the iSEE package to be essential for exploring and showcasing relevant features of any data set that can be represented in the form of a SummarizedExperiment object.
-In particular, we expect public instances of iSEE to ideally accompany publications, as they enable authors to reproduce published figures and showcase salient aspects of their work using guided tours, while fostering hypothesis generation through transparent, interactive, reproducible, and efficient data analysis by the broader reader audience.
+iSEE provides a general interactive interface for visual exploration of high-throughput biological data sets.
+Any study that can be represented in a SummarizedExperiment object can be used as input, allowing iSEE to accommodate a diverse range of `-omics data sets.
+The interface is flexible and can be dynamically customized by the user; supports exploration of interactions between data aspects through colouring and linking between panels;
+and provides transparency and reproducibility during the interactive analysis, through code tracking and state reporting. 
+The most obvious use of iSEE is that of data exploration for hypothesis generation during the course of a research project.
+However, we also anticipate that public instances of iSEE will accompany publications to enable authors to showcase salient aspects of their work using guided tours.
 
-\subsection*{Author contributions}
+\section*{Data and software availability}
+The iSEE package is available at \url{https://bioconductor.org/packages/iSEE}.
+Source code of the development version of the package is available at \url{https://github.com/csoneson/iSEE}.
+Code for the demonstrations and tours is available at \url{https://github.com/LTLA/iSEE2018}.
+
+\section*{Author contributions}
 All authors contributed conceptually and technically to the development and maintenance of large sections of the package code.
@@
 
-\subsection*{Data and software availability}
-The iSEE package is available at \url{https://bioconductor.org/packages/iSEE}
+\section{Competing interests}
+No competing interests were declared.
 
-Source code of the development version is available at \url{https://github.com/csoneson/iSEE}
-
-Archived source code as at the time of publication is available at % TODO
-
-An instance of the package running on a public Shiny Server is available at \url{http://shiny.imbei.uni-mainz.de:3838/iSEE/}
-
-% TODO: docker/singularity containers
-
-% TODO:conda pkg
-% The iSEE package is also distributed via the bioconda channel for the conda package manager at...https://anaconda.org/bioconda/bioconductor-isee % still needs to be finalized!!
-
-Software license: MIT
-
-\subsection*{Competing interests}
-All financial, personal, or professional competing interests for any of the authors that
-could be construed to unduly influence the content of the article must be disclosed and
-will be displayed alongside the article.
-
-\subsection*{Grant information}
+\section{Grant information}
 ATLL was supported by core funding from Cancer Research UK (award no. 17197 to Dr.\ John Marioni).
@@
 
-\subsection*{Acknowledgements}
+\section{Acknowledgements}
 We thank the organizers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.

### 2018-04-30 manuscript/main.tex

@@
 All authors contributed conceptually and technically to the development and maintenance of large sections of the package code.
-A.T.L.L. and F.M. developed the application and data infrastructure.
-K.R.A and C.S. developed the plotting framework.
-F.M. developed the code tracking and user interface.
-K.R.A designed and developed the ExperimentColorMap class.
+ATLL and FM developed the application and data infrastructure.
+KRA and CS developed the plotting framework.
+FM developed the code tracking and user interface.
+KRA designed and developed the ExperimentColorMap class.
 All authors contributed to testing the application on all data sets.

### 2018-04-30 manuscript/main.tex

@@
 
-\section{Competing interests}
+\section*{Competing interests}
 No competing interests were declared.
 
-\section{Grant information}
+\section*{Grant information}
 ATLL was supported by core funding from Cancer Research UK (award no. 17197 to Dr.\ John Marioni).
@@
 
-\section{Acknowledgements}
+\section*{Acknowledgements}
 We thank the organizers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.

### 2018-04-30 tours/cytof_app.R
edited cytof use case: split into data+app, fixed the call for the server

@@ -1,6 +1,5 @@
-## Read data object and tour steps
+## Read tour steps
 tour_steps <- read.delim("cytof_tour.txt", header = TRUE, as.is = TRUE, sep = ";")
-load(url("https://www.dropbox.com/s/ogqp5grxe39c02d/bcrxl_t.rda?dl=1"))  ## bcrxl object
 
 ## Packages
 stopifnot(require(iSEE))
@@ -42,14 +41,12 @@ fd$BrushData <- list(NULL, NULL, NULL,
                      list(xmin = -1.3837629808554, xmax = 1.6091336966254, ymin = 0.38148152957448, ymax = 7.4740835739801, mapping = list(x = "X", y = "Y"), domain = list(left = -3.80032310964924, right = 9.1168276366504, bottom = 0.248462505430422, top = 7.95803222728227), range = list(left = 32.5671687714041, right = 209.520547945205, bottom = 374.038986186765, top = 24.0290131340161), log = list(x = NULL, y = NULL), direction = "xy", brushId = "featExprPlot5_Brush", outputId = "featExprPlot5"),
                      NULL)
 
-app <- iSEE(bcrxl, tour = tour_steps,
-            redDimArgs = NULL, colDataArgs = NULL,
-            featExprArgs = fd,
-            rowStatArgs = NULL, rowDataArgs = NULL, heatMapArgs = NULL,
-            redDimMax = 0, colDataMax = 0,
-            featExprMax = nrow(fd),
-            rowStatMax = 0, rowDataMax = 0, heatMapMax = 0,
-            initialPanels = initialPanels,
-            appTitle = "CyTOF data tour")
-
-shiny::runApp(app)
+iSEE(bcrxl, tour = tour_steps,
+     redDimArgs = NULL, colDataArgs = NULL,
+     featExprArgs = fd,
+     rowStatArgs = NULL, rowDataArgs = NULL, heatMapArgs = NULL,
+     redDimMax = 0, colDataMax = 0,
+     featExprMax = nrow(fd),
+     rowStatMax = 0, rowDataMax = 0, heatMapMax = 0,
+     initialPanels = initialPanels,
+     appTitle = "CyTOF data tour")

### 2018-04-30 tours/cytof_data.R

@@ -1,0 +1,2 @@
+## Read data object
+load(url("https://www.dropbox.com/s/ogqp5grxe39c02d/bcrxl_t.rda?dl=1"))  ## bcrxl object

### 2018-05-02 tours/cytof_app.R
renamed the cytof part

@@ -4,11 +4,11 @@ tour_steps <- read.delim("cytof_tour.txt", header = TRUE, as.is = TRUE, sep = ";
 ## Packages
 stopifnot(require(iSEE))
 
-## Set up panels (6 feature expression plots)
-fd <- featExprPlotDefaults(bcrxl, 6)
+## Set up panels (6 feature assay plots)
+fd <- featAssayPlotDefaults(bcrxl, 6)
 
 initialPanels = DataFrame(
-    Name = c(sprintf("Feature expression plot %i", seq(1, nrow(fd), 1))),
+    Name = c(sprintf("Feature assay plot %i", seq(1, nrow(fd), 1))),
     Width = 4)
 
 ## Appearance
@@ -34,19 +34,19 @@ fd$Downsample <- TRUE
 fd$SampleRes <- 100
 
 ## Point selection
-fd$SelectByPlot <- c("", "", "", "", "Feature expression plot 4", "Feature expression plot 5")
+fd$SelectByPlot <- c("", "", "", "", "Feature assay plot 4", "Feature assay plot 5")
 fd$SelectEffect <- "Restrict"
 fd$BrushData <- list(NULL, NULL, NULL,
-                     list(xmin = -3.2643893296676, xmax = 6.2505437028317, ymin = 3.9638278603138, ymax = 6.6394433908566, mapping = list(x = "X", y = "Y"), domain = list(left = -3.77917836650223, right = 8.80210985412216, bottom = -3.47327157913897, top = 7.12045292443739), range = list(left = 44.2370906464041, right = 209.520547945205, bottom = 374.038986186765, top = 21.6559662590161), log = list(x = NULL, y = NULL), direction = "xy", brushId = "featExprPlot4_Brush", outputId = "featExprPlot4"),
-                     list(xmin = -1.3837629808554, xmax = 1.6091336966254, ymin = 0.38148152957448, ymax = 7.4740835739801, mapping = list(x = "X", y = "Y"), domain = list(left = -3.80032310964924, right = 9.1168276366504, bottom = 0.248462505430422, top = 7.95803222728227), range = list(left = 32.5671687714041, right = 209.520547945205, bottom = 374.038986186765, top = 24.0290131340161), log = list(x = NULL, y = NULL), direction = "xy", brushId = "featExprPlot5_Brush", outputId = "featExprPlot5"),
[…]

### 2018-05-02 tours/pbmc4k_app.R
renamed the pbmc4k part

@@ -8,7 +8,7 @@ sce <- readRDS("sce.rds")
 # Setting up defaults for all panels.
 redDim <- DataFrame(Type="TSNE", ColorBy="Column data", ColorByColData="Detection")
 colData <- DataFrame(YAxis="log10_total_counts", XAxis="Column data", XAxisColData="Cluster", ColorBy="Column data", ColorByColData="Detection")
-featExpr <- DataFrame(YAxisFeatName="PF4", XAxis="Column data", XAxisColData="Cluster", ColorBy="Column data", ColorByColData="log10_total_counts")
+featAssay <- DataFrame(YAxisFeatName="PF4", XAxis="Column data", XAxisColData="Cluster", ColorBy="Column data", ColorByColData="log10_total_counts")
 heatMap <- DataFrame(FeatName=I(list(c("PF4", "PPBP", "TMSB4X", "B2M", "ACTB", "SDPR"))), 
                      ColData=I(list(c("Cluster"))), Lower=-5, Upper=5)
 rowStat <- DataFrame(Search="PF4", Selected="PF4")
@@ -18,10 +18,10 @@ colData$SelectByPlot <- "Reduced dimension plot 1"
 colData$SelectEffect <- "Transparent"
 colData$SelectAlpha <- 0
 
-featExpr$SelectByPlot <- "Reduced dimension plot 1"
-featExpr$SelectEffect <- "Transparent"
-featExpr$SelectAlpha <- 0
-featExpr$YAxisRowTable <- "Row statistics table 1"
+featAssay$SelectByPlot <- "Reduced dimension plot 1"
+featAssay$SelectEffect <- "Transparent"
+featAssay$SelectAlpha <- 0
+featAssay$YAxisRowTable <- "Row statistics table 1"
 
 heatMap$SelectByPlot <- "Reduced dimension plot 1"
 
@@ -32,7 +32,7 @@ annot.fun <- annotateEnsembl(sce, orgdb=org.Hs.eg.db, keytype="ENSEMBL", rowdata
 # Setting up the initial panels.
 init <- DataFrame(Name=c("Reduced dimension plot 1",
                          "Column data plot 1",
-                         "Feature expression plot 1",
+                         "Feature assay plot 1",
                          "Row statistics table 1",
                          "Heat map 1"),
                   Width=c(4,4,4,8,4))
[…]

### 2018-05-02 tours/tcga_app.R
renamed the tcga part

@@ -93,8 +93,8 @@ rd$LegendPosition <- "Right"
 rd$SelectBoxOpen <- TRUE
 rd$SelectByPlot <- c("Column data plot 1", "Column data plot 2")
 
-# Panel 3: feature expression (analysis)
-fe <- featExprPlotDefaults(sce, 2)
+# Panel 3: feature assay (analysis)
+fe <- featAssayPlotDefaults(sce, 2)
 # data
 fe$DataBoxOpen <- TRUE
 fe$Assay <- 2
@@ -117,12 +117,12 @@ initialPanels = DataFrame(
     Name = c(
       "Column data plot 1",
       "Reduced dimension plot 1",
-      "Feature expression plot 1"
+      "Feature assay plot 1"
     ),
     Width = c(
       3, # colData
       4, # reducedDim
-      5 # featExprs
+      5 # featAssays
     )
   )
 
@@ -130,9 +130,9 @@ tour <- read.delim(tour_steps, sep=";", stringsAsFactors = FALSE,row.names = NUL
 
 app <- iSEE(
   sce, tour = tour,
-  redDimArgs = rd, colDataArgs = cd, featExprArgs = fe,
+  redDimArgs = rd, colDataArgs = cd, featAssayArgs = fe,
   rowStatArgs = NULL, rowDataArgs = NULL, heatMapArgs = NULL,
-  redDimMax = 1, colDataMax = 1, featExprMax = 1,
+  redDimMax = 1, colDataMax = 1, featAssayMax = 1,
   rowStatMax = 1, rowDataMax = 1, heatMapMax = 1,
   initialPanels = initialPanels,
   appTitle = "TCGA RNA-seq tour")

### 2018-05-02 manuscript/pics/setup.R
renamed done in the pics setup for the MS

@@ -38,15 +38,15 @@ redDim <- DataFrame(ColorBy="Feature name", ColorByFeatName="Cux2",
 colData <- DataFrame(YAxis="NALIGNED", XAxis="Column data", XAxisColData="PRIMER",
                      ColorBy="Column data", ColorByColData="passes_qc_checks_s",
 					 FontSize=1.5, PointSize=1.5)
-featExpr <- DataFrame(YAxisFeatName="Cux2", ColorBy="Column data", ColorByColData="driver_1_s",
+featAssay <- DataFrame(YAxisFeatName="Cux2", ColorBy="Column data", ColorByColData="driver_1_s",
 					 FontSize=1.5, PointSize=1.5)
 heatMap <- DataFrame(FeatName=I(list(c("Fam19a1", "Cacna2d3", "Cux2", "Rorb", "Ptn", "Lmo3", "Foxp2", "Lamp5", "Etv1"))),
                      ColData=I(list(c("driver_1_s"))))
 rowStat <- DataFrame(Selected="Cux2", Search="homeobox 2")
 
-init <- data.frame(Name=c("Reduced dimension plot 1", "Column data plot 1", "Feature expression plot 1",
+init <- data.frame(Name=c("Reduced dimension plot 1", "Column data plot 1", "Feature assay plot 1",
                           "Row statistics table 1", "Heat map 1"),
                    Width=c(4,4,4,8,4))
 
 library(iSEE)
-iSEE(sce, redDimArgs=redDim, colDataArgs=colData, featExprArgs=featExpr, heatMapArgs=heatMap, rowDataMax=0, rowStatArgs=rowStat, initialPanels=init)
+iSEE(sce, redDimArgs=redDim, colDataArgs=colData, featAssayArgs=featAssay, heatMapArgs=heatMap, rowDataMax=0, rowStatArgs=rowStat, initialPanels=init)

### 2018-05-02 tours/tcga_app.R
move TCGA data-related preprocessing to _data.R

@@ -1,38 +1,19 @@
 
-# User inputs ----
+# Tour steps ----
 
 tour_steps <- "tcga_tour.txt"
 stopifnot(file.exists(tour_steps))
+tour <- read.delim(tour_steps, sep=";", stringsAsFactors = FALSE,row.names = NULL)
 
 # Libraries ----
 
 stopifnot(require(iSEE))
-stopifnot(requireNamespace("ExperimentHub"))
-stopifnot(requireNamespace("scater"))
-stopifnot(requireNamespace("irlba"))
-stopifnot(requireNamespace("Rtsne"))
-
-# Preprocessing ----
-
-# Vignette demo
-
-ehub <- ExperimentHub::ExperimentHub()
-eh1 <- ehub[["EH1"]] # an ExpressionSet
-se1 <- as(eh1, "SummarizedExperiment")
-sce <- as(se1, "SingleCellExperiment")
-sce <- scater::runPCA(sce, exprs_values = "exprs")
-irlba_out <- irlba::irlba(assay(sce, "exprs"))
-tsne_out <- Rtsne::Rtsne(
-  irlba_out$v, pca = FALSE, perplexity = 50, verbose = TRUE)
-reducedDim(sce, "TSNE") <- tsne_out$Y
-
-# library size and CPM
-
-colData(sce)[,"lib_size"] <- colSums(assay(sce, "exprs"))
-assay(sce, "log2CPM") <-
-  log2(1 + t(t(assay(sce, "exprs")) / colData(sce)[,"lib_size"]))
 
 # Application setup ----
+
+if (length(ls(pattern = "sce")) != 1){
+  stop("Please run tcga_data.R first to generate the `sce` object.")
+}
 
 # Panel 1: colData (phenotype selection)
 # Y = CancerType
@@ -125,8 +106,6 @@ initialPanels = DataFrame(
       5 # featAssays
     )
   )
-
-tour <- read.delim(tour_steps, sep=";", stringsAsFactors = FALSE,row.names = NULL)
 
 app <- iSEE(
   sce, tour = tour,

### 2018-05-02 tours/tcga_data.R

@@ -1,0 +1,28 @@
+
+# Librarires ----
+
+stopifnot(requireNamespace("ExperimentHub"))
+stopifnot(require(SingleCellExperiment))
+stopifnot(requireNamespace("scater"))
+stopifnot(requireNamespace("irlba"))
+stopifnot(requireNamespace("Rtsne"))
+
+# Preprocessing ----
+
+# Vignette demo
+
+ehub <- ExperimentHub::ExperimentHub()
+eh1 <- ehub[["EH1"]] # an ExpressionSet
+se1 <- as(eh1, "SummarizedExperiment")
+sce <- as(se1, "SingleCellExperiment")
+sce <- scater::runPCA(sce, exprs_values = "exprs")
+irlba_out <- irlba::irlba(assay(sce, "exprs"))
+tsne_out <- Rtsne::Rtsne(
+  irlba_out$v, pca = FALSE, perplexity = 50, verbose = TRUE, max_iter = 100)
+reducedDim(sce, "TSNE") <- tsne_out$Y
+
+# library size and CPM
+
+colData(sce)[,"lib_size"] <- colSums(assay(sce, "exprs"))
+assay(sce, "log2CPM") <-
+  log2(1 + t(t(assay(sce, "exprs")) / colData(sce)[,"lib_size"]))

### 2018-05-02 tours/tcga_data.R
testing leftover

@@ -18,7 +18,7 @@ sce <- as(se1, "SingleCellExperiment")
 sce <- scater::runPCA(sce, exprs_values = "exprs")
 irlba_out <- irlba::irlba(assay(sce, "exprs"))
 tsne_out <- Rtsne::Rtsne(
-  irlba_out$v, pca = FALSE, perplexity = 50, verbose = TRUE, max_iter = 100)
+  irlba_out$v, pca = FALSE, perplexity = 50, verbose = TRUE)
 reducedDim(sce, "TSNE") <- tsne_out$Y
 
 # library size and CPM

### 2018-05-05 manuscript/main.tex
Correct affiliations CS

@@
 \author[2,3,$\dagger$]{Federico Marini}
-\author[4,$\dagger$]{Charlotte Soneson}
-\author[5,$\dagger$,$\ast$]{Aaron T. L. Lun}
+\author[4,5,$\dagger$]{Charlotte Soneson}
+\author[6,$\dagger$,$\ast$]{Aaron T. L. Lun}
 
@@
 \affil[3]{Institute for Medical Biostatistics, Epidemiology and Informatics (IMBEI), University Medical Center of the Johannes Gutenberg University Mainz, Mainz}
-\affil[4]{University of Zurich and SIB Swiss Institute of Bioinformatics, Zurich, Switzerland}
-\affil[5]{Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom}
+\affil[4]{Institute of Molecular Life Sciences, University of Zurich, 8057 Zurich, Switzerland}
+\affil[5]{SIB Swiss Institute of Bioinformatics, University of Zurich, 8057 Zurich, Switzerland}
+\affil[6]{Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom}
 \affil[$\dagger$]{These authors contributed equally.}
@@
 \item Column data plots, for visualizing sample metadata stored in the colData slot of SummarizedExperiment object.
-\item Feature expression plots, for visualizing expression values for a particular feature (e.g., gene) across samples from any assay in the SummarizedExperiment object.
+\item Feature assay plots, for visualizing expression values for a particular feature (e.g., gene) across samples from any assay in the SummarizedExperiment object.
 Despite its name, this plot type is applicable to data modalities beyond gene expression.
@@
 (B) Information can be transmitted between panels according to a user-specified scheme.
-Here, the selection of feature $X$ in the row statistics table determines the y-axis of the feature expression plot, and colours the samples in the reduced dimension plot by the expression of $X$.
+Here, the selection of feature $X$ in the row statistics table determines the y-axis of the feature assay plot, and colours the samples in the reduced dimension plot by the expression of $X$.
 Selection of points in the reduced dimension plot (dotted blue line) also determines the samples that are shown in the column data (i.e., sample metadata) plot;
@@
 
-Each sample is represented as a point in column data, feature expression and reduced dimension plots.
+Each sample is represented as a point in column data, feature assay and reduced dimension plots.
 Similarly, each feature is represented by a point in row data plots.
@@
 This feature facilitates exploration of the relationships between different aspects of the data.
-For example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding -- this is achieved by selecting points in a reduced dimension plot (using the standard rectangular brush or a lasso) and transmitting that selection to any number of feature expression plots.\\
+For example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding -- this is achieved by selecting points in a reduced dimension plot (using the standard rectangular brush or a lasso) and transmitting that selection to any number of feature assay plots.\\
 

### 2018-05-05 manuscript/main.tex
Some tweaks to the text

@@
 \section*{Introduction}
-Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as genomics.
+Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays commonly used in genomics.
 Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
-To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).\\
+To this end, visualization of the data in an intuitive and interactive interface is crucial for allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).\\
 
@@
 Each instance of the SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where rows and columns represent genomic features and biological samples, respectively.
-For instance, individual assays may represent gene expression matrices of raw counts along with different normalization strategies.
+For instance, individual assays may represent gene expression matrices, either in the form of raw counts or as values obtained by different normalization strategies.
 In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; those may include experimental metadata as well as any analysis results.\\
@@
 \begin{itemize}
-\item Column data plots, for visualizing sample metadata stored in the colData slot of SummarizedExperiment object.
+\item Column data plots, for visualizing sample metadata stored in the colData slot of the SummarizedExperiment object.
 \item Feature assay plots, for visualizing expression values for a particular feature (e.g., gene) across samples from any assay in the SummarizedExperiment object.
@@
 \item Row statistics tables, to present the contents of the rowData slot for the SummarizedExperiment object.
-\item Row data plots, for visualizing sample metadata stored in the rowData slot of a SummarizedExperiment object.
+\item Row data plots, for visualizing feature metadata stored in the rowData slot of the SummarizedExperiment object.
 \item Heat maps, to visualize assay data for multiple features where samples are ordered by one or more colData fields.
 \item Reduced dimension plots, which display any two dimensions from pre-computed dimensionality reduction results (e.g., from PCA or $t$-SNE).
-These results are taken from the reducedDim slot of any SingleCellExperiment object supplied to iSEE.
+These results are taken from the reducedDim slot if this is available in the object supplied to iSEE.
 \end{itemize}
@@
 Panel types are available to visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
-Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
+Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heat maps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
 Panels of each type are colour-coded for ease of interpretation.
@@
 Selection of points in the reduced dimension plot (dotted blue line) also determines the samples that are shown in the column data (i.e., sample metadata) plot;
-further selection of points in the column data plot determines the samples that are shown in the heatmap.
+further selection of points in the column data plot determines the samples that are shown in the heat map.
 }
@@
 For these panel types, a scatter plot is automatically produced if the selected variables on the x- and y-axes are both continuous.
-If exactly one variable is categorical, points are grouped by the categorical levels and a (horizontal) violin plot is produced with points scattered within each violin.
+If exactly one variable is categorical, points are grouped by the categorical levels and a (vertical or horizontal) violin plot is produced with points scattered within each violin.
 If both variables are categorical, a ``rectangle plot'' is produced where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of that combination.
@@
 \subsection*{Custom panel colouring}
-Sample-based points can be coloured according to the values of any sample-level metadata field in the colData or by the assay values of a selected feature.
-Similarly, feature-based points can be coloured according to any feature-level metadata field in the rowData.
-Heatmaps are coloured according to the expression values of the selected features in the chosen assay,
-with additional colouring of each of the colData fields used to order the samples.
+Sample-based points can be coloured according to the values of any sample-level metadata field in the colData slot or by the assay values of a selected feature.
+Similarly, feature-based points can be coloured according to any feature-level metadata field in the rowData slot.
+Heat maps are coloured according to the expression values of the selected features in the chosen assay,
+with additional colour annotation by each of the colData fields used to order the samples.
 In all cases, the variable to use for colouring can be dynamically selected for each plot.
@@
 This feature facilitates exploration of the relationships between different aspects of the data.
-For example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding -- this is achieved by selecting points in a reduced dimension plot (using the standard rectangular brush or a lasso) and transmitting that selection to any number of feature assay plots.\\
+For example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding -- this is achieved by selecting points in a reduced dimension plot (using the standard rectangular brush or a lasso selection) and transmitting that selection to any number of feature assay plots.\\
 
@@
 Row statistics tables can also transmit to various plot types, by selecting a table row to control the coloring of sample-based points;
-or by defining a subset of features to visualize in a heatmap.
+or by defining a subset of features to visualize in a heat map.
 Furthermore, row data plots can transmit to row statistics tables, whereby selection of points in the former will subset the latter.
@@
 
-iSEE automatically memorizes the exact R code that was used to generate every plot, extending previous work by \cite{marini2016interrepro}.
+iSEE automatically memorizes the exact R code that was used to generate every plot, extending previous work by \citet{marini2016interrepro}.
 This code is fully accessible to users at any time during the run-time of the interface.
@@
 \subsection*{Additional functionalities}
-Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez.
+Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl \citep{zerbino2018ensembl} or Entrez \citep{ncbi2017entrez}.
 % do we need to cite them?
@@
 iSEE provides a general interactive interface for visual exploration of high-throughput biological data sets.
-Any study that can be represented in a SummarizedExperiment object can be used as input, allowing iSEE to accommodate a diverse range of `-omics data sets.
+Any study that can be represented in a SummarizedExperiment object can be used as input, allowing iSEE to accommodate a diverse range of `omics data sets.
 The interface is flexible and can be dynamically customized by the user; supports exploration of interactions between data aspects through colouring and linking between panels;

### 2018-05-05 manuscript/main.tex
heat map -> heatmap, remove sentence

@@
 \item Feature assay plots, for visualizing expression values for a particular feature (e.g., gene) across samples from any assay in the SummarizedExperiment object.
-Despite its name, this plot type is applicable to data modalities beyond gene expression.
 \item Row statistics tables, to present the contents of the rowData slot for the SummarizedExperiment object.
 \item Row data plots, for visualizing feature metadata stored in the rowData slot of the SummarizedExperiment object.
-\item Heat maps, to visualize assay data for multiple features where samples are ordered by one or more colData fields.
+\item Heatmaps, to visualize assay data for multiple features where samples are ordered by one or more colData fields.
 \item Reduced dimension plots, which display any two dimensions from pre-computed dimensionality reduction results (e.g., from PCA or $t$-SNE).
@@
 Panel types are available to visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
-Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heat maps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
+Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
 Panels of each type are colour-coded for ease of interpretation.
@@
 Selection of points in the reduced dimension plot (dotted blue line) also determines the samples that are shown in the column data (i.e., sample metadata) plot;
-further selection of points in the column data plot determines the samples that are shown in the heat map.
+further selection of points in the column data plot determines the samples that are shown in the heatmap.
 }
@@
 Similarly, feature-based points can be coloured according to any feature-level metadata field in the rowData slot.
-Heat maps are coloured according to the expression values of the selected features in the chosen assay,
+Heatmaps are coloured according to the expression values of the selected features in the chosen assay,
 with additional colour annotation by each of the colData fields used to order the samples.
@@
 Row statistics tables can also transmit to various plot types, by selecting a table row to control the coloring of sample-based points;
-or by defining a subset of features to visualize in a heat map.
+or by defining a subset of features to visualize in a heatmap.
 Furthermore, row data plots can transmit to row statistics tables, whereby selection of points in the former will subset the latter.

### 2018-05-05 manuscript/pics/setup.R
Updated screenshot setup script.

@@ -37,16 +37,16 @@ redDim <- DataFrame(ColorBy="Feature name", ColorByFeatName="Cux2",
                     FontSize=1.5, PointSize=1.5)
 colData <- DataFrame(YAxis="NALIGNED", XAxis="Column data", XAxisColData="PRIMER",
                      ColorBy="Column data", ColorByColData="passes_qc_checks_s",
-					 FontSize=1.5, PointSize=1.5)
-featExpr <- DataFrame(YAxisFeatName="Cux2", ColorBy="Column data", ColorByColData="driver_1_s",
-					 FontSize=1.5, PointSize=1.5)
+                     FontSize=1.5, PointSize=1.5)
+featAssay <- DataFrame(YAxisFeatName="Cux2", ColorBy="Column data", ColorByColData="driver_1_s",
+                       FontSize=1.5, PointSize=1.5)
 heatMap <- DataFrame(FeatName=I(list(c("Fam19a1", "Cacna2d3", "Cux2", "Rorb", "Ptn", "Lmo3", "Foxp2", "Lamp5", "Etv1"))),
                      ColData=I(list(c("driver_1_s"))))
 rowStat <- DataFrame(Selected="Cux2", Search="homeobox 2")
 
-init <- data.frame(Name=c("Reduced dimension plot 1", "Column data plot 1", "Feature expression plot 1",
+init <- data.frame(Name=c("Reduced dimension plot 1", "Column data plot 1", "Feature assay plot 1",
                           "Row statistics table 1", "Heat map 1"),
                    Width=c(4,4,4,8,4))
 
 library(iSEE)
-iSEE(sce, redDimArgs=redDim, colDataArgs=colData, featExprArgs=featExpr, heatMapArgs=heatMap, rowDataMax=0, rowStatArgs=rowStat, initialPanels=init)
+iSEE(sce, redDimArgs=redDim, colDataArgs=colData, featAssayArgs=featAssay, heatMapArgs=heatMap, rowDataMax=0, rowStatArgs=rowStat, initialPanels=init)

### 2018-05-05 manuscript/main.tex

--- 2018-05-05 manuscript/main.tex
@@
 \author[2,3,$\dagger$]{Federico Marini}
-\author[4,5,$\dagger$]{Charlotte Soneson}
-\author[6,$\dagger$,$\ast$]{Aaron T. L. Lun}
+\author[4,$\dagger$]{Charlotte Soneson}
+\author[5,$\dagger$,$\ast$]{Aaron T. L. Lun}
 
@@
 \affil[3]{Institute for Medical Biostatistics, Epidemiology and Informatics (IMBEI), University Medical Center of the Johannes Gutenberg University Mainz, Mainz}
-\affil[4]{Institute of Molecular Life Sciences, University of Zurich, 8057 Zurich, Switzerland}
-\affil[5]{SIB Swiss Institute of Bioinformatics, University of Zurich, 8057 Zurich, Switzerland}
-\affil[6]{Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom}
+\affil[4]{University of Zurich and SIB Swiss Institute of Bioinformatics, Zurich, Switzerland}
+\affil[5]{Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom}
 \affil[$\dagger$]{These authors contributed equally.}

### 2018-05-08 manuscript/main.tex

--- 2018-05-05 manuscript/main.tex
@@
 Data exploration is critical to the comprehension of large biological data sets generated by high-throughput assays such as sequencing.
-However, most existing tools for interactive visualization are limited to specific assays or analyses.
+However, most existing tools for interactive visualisation are limited to specific assays or analyses.
 Here, we present the iSEE (Interactive SummarizedExperiment Explorer) software package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
-iSEE is directly compatible with many existing R/Bioconductor packages for analyzing high-throughput biological data,
+iSEE is directly compatible with many existing R/Bioconductor packages for analysing high-throughput biological data,
 and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
@@
 
-Single-cell, gene expression, transcriptomics, proteomics, visualization, interactive, shiny, Bioconductor
+Single-cell, gene expression, transcriptomics, proteomics, visualisation, interactive, shiny, Bioconductor
 
@@
 Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
-To this end, visualization of the data in an intuitive and interactive interface is crucial for allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).\\
+To this end, visualisation of the data in an intuitive and interactive interface is crucial for allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).\\
 
-Most existing tools for interactive visualization of biological data are designed for specific assays and analyses, e.g., pRoloc for proteomics \citep{gatto2014mass}, shinyMethyl for methylation \citep{fortin2014shinymethyl}, HTSvis for high-throughput screens \citep{scheeder2017htsvis}.
-Opportunities for customization are generally limited, making it difficult to re-use the same visualization software for new technologies or experimental designs where different aspects of the data are of interest.
-Moreover, standalone tools such as the Loupe Cell Browser from 10X Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}.
+Most existing tools for interactive visualisation of biological data are designed for specific assays and analyses, e.g., pRoloc for proteomics \citep{gatto2014mass}, shinyMethyl for methylation \citep{fortin2014shinymethyl}, HTSvis for high-throughput screens \citep{scheeder2017htsvis}.
+Opportunities for customisation are generally limited, making it difficult to re-use the same visualisation software for new technologies or experimental designs where different aspects of the data are of interest.
+Moreover, standalone tools such as the Loupe Cell Browser from 10x Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}.
 This complicates any coordinated use of these tools with a reproducible, transparent, and statistically rigorous analysis.\\
@@
 iSEE is implemented in R using the Shiny framework \citep{chang2017shiny} and exploits data structures from the open-source Bioconductor project \citep{gentleman2004bioconductor}, specifically the SummarizedExperiment class.
-iSEE allows users to simultaneously visualize multiple aspects of a given data set, including experimental data, metadata and analysis results.
+iSEE allows users to simultaneously visualise multiple aspects of a given data set, including experimental data, metadata and analysis results.
 Dynamic linking and point selection facilitate the flexible exploration of interactions between different data aspects.
@@
 \section*{Taking a SummarizedExperiment input}
-The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
+The iSEE interface is initialised with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
 Each instance of the SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where rows and columns represent genomic features and biological samples, respectively.
-For instance, individual assays may represent gene expression matrices, either in the form of raw counts or as values obtained by different normalization strategies.
-In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; those may include experimental metadata as well as any analysis results.\\
+For instance, individual assays may represent gene expression matrices, either in the form of raw counts or as values obtained by different normalisation strategies.
+In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'' slots, respectively; those may include experimental metadata as well as any analysis results.\\
 
@@
 Package developers can also easily use the base SummarizedExperiment class to derive new bespoke classes for particular applications, such as the SingleCellExperiment class for single-cell 'omics data.
-By accepting SummarizedExperiment objects as input, iSEE immediately offers interactive visualization for a variety of data modalities.
+By accepting SummarizedExperiment objects as input, iSEE immediately offers interactive visualisation for a variety of data modalities.
 This complements the state-of-the-art analysis workflows and methodologies already available in R/Bioconductor packages. \\
@@
 All data aspects stored in a SummarizedExperiment can be readily examined alongside one another in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
-The interface layout is built using the shinydashboard package \citep{chang2018shinydashboard}, using colour-coded panels to visualize each data aspect.
+The interface layout is built using the shinydashboard package \citep{chang2018shinydashboard}, using colour-coded panels to visualise each data aspect.
 Individual panel types include:
@@
 \begin{itemize}
-\item Column data plots, for visualizing sample metadata stored in the colData slot of the SummarizedExperiment object.
-\item Feature assay plots, for visualizing expression values for a particular feature (e.g., gene) across samples from any assay in the SummarizedExperiment object.
+\item Column data plots, for visualising sample metadata stored in the colData slot of the SummarizedExperiment object.
+\item Feature assay plots, for visualising expression values for a particular feature (e.g., gene) across samples from any assay in the SummarizedExperiment object.
 \item Row statistics tables, to present the contents of the rowData slot for the SummarizedExperiment object.
-\item Row data plots, for visualizing feature metadata stored in the rowData slot of the SummarizedExperiment object.
-\item Heatmaps, to visualize assay data for multiple features where samples are ordered by one or more colData fields.
+\item Row data plots, for visualising feature metadata stored in the rowData slot of the SummarizedExperiment object.
+\item Heatmaps, to visualise assay data for multiple features where samples are ordered by one or more colData fields.
 \item Reduced dimension plots, which display any two dimensions from pre-computed dimensionality reduction results (e.g., from PCA or $t$-SNE).
@@
 \includegraphics[width=\textwidth]{pics/Fig1.pdf}
-\caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data.
+\caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualises a different aspect of the data.
 New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
-Panel types are available to visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
-Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
+Panel types are available to visualise sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
+Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualise experimental observations for multiple features; and feature-level metadata plots.
 Panels of each type are colour-coded for ease of interpretation.
@@
 For these panel types, a scatter plot is automatically produced if the selected variables on the x- and y-axes are both continuous.
-If exactly one variable is categorical, points are grouped by the categorical levels and a (vertical or horizontal) violin plot is produced with points scattered within each violin.
+If exactly one variable is categorical, points are grouped by the categorical levels and a (vertical or horisontal) violin plot is produced with points scattered within each violin.
 If both variables are categorical, a ``rectangle plot'' is produced where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of that combination.
@@
 The returned colours will be mapped to factor levels for categorical variables, or used in colour interpolation for continuous variables.
-For categorical variables, the function may also return a constant vector of named colors corresponding to the levels of a known factor.
+For categorical variables, the function may also return a constant vector of named colours corresponding to the levels of a known factor.
 Colour maps can be specified for individual variables; for all assays, all column data variables, or all row data variables (with different functions for continuous or categorical variables); or for all categorical or continuous variables.
-This provides a convenient yet flexible mechanism for customization of colouring schemes within the interface.
+This provides a convenient yet flexible mechanism for customisation of colouring schemes within the interface.
 
@@
 A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
-Users can define and reorganize arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting panels control the inclusion and appearance of the corresponding data points in receiving panels.
+Users can define and reorganise arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting panels control the inclusion and appearance of the corresponding data points in receiving panels.
 This feature facilitates exploration of the relationships between different aspects of the data.
@@
 With iSEE, this concept is extended to any assay data, feature-level or sample-level metadata present in a SummarizedExperiment object, providing a powerful framework for interrogating multiple interactions between data aspects.
-Row statistics tables can also transmit to various plot types, by selecting a table row to control the coloring of sample-based points;
-or by defining a subset of features to visualize in a heatmap.
+Row statistics tables can also transmit to various plot types, by selecting a table row to control the colouring of sample-based points;
+or by defining a subset of features to visualise in a heatmap.
 Furthermore, row data plots can transmit to row statistics tables, whereby selection of points in the former will subset the latter.
@@
 
-iSEE automatically memorizes the exact R code that was used to generate every plot, extending previous work by \citet{marini2016interrepro}.
+iSEE automatically memorises the exact R code that was used to generate every plot, extending previous work by \citet{marini2016interrepro}.
 This code is fully accessible to users at any time during the run-time of the interface.
@@
 Similarly, the code required to reproduce the current state of the interface can also be reported.
-This can be used in startup scripts to launch an iSEE instance in any preferred layout, including the panel organization, variable selection, colouring schemes, links between panels and even individual brushes and lasso selections.
+This can be used in startup scripts to launch an iSEE instance in any preferred layout, including the panel organisation, variable selection, colouring schemes, links between panels and even individual brushes and lasso selections.
 
@@
 at least ENSEMBL has updated citable papers
-For large data sets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
+For large data sets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualisation.
 Users can also include a bespoke step-by-step ``tour'' of their data set via the rintrojs package \cite{ganz2016rintrojs}, guiding the audience through an examination of the salient features in the data.
@@
 \subsection*{Droplet-based single-cell RNA sequencing}
-We applied iSEE to a larger scRNA-seq data set involving 4,000 peripheral blood mononuclear cells (PBMCs), generated by 10X Genomics \citep{zheng2017massively}.
+We applied iSEE to a larger scRNA-seq data set involving 4,000 peripheral blood mononuclear cells (PBMCs), generated by 10x Genomics \citep{zheng2017massively}.
 This demonstration explores the differences between different methods for distinguishing cells from empty droplets in droplet-based scRNA-seq protocols \citep{lun2018distinguishing}.\\
@@
 \subsection*{Mass cytometry}
-Finally, we explored a mass cytometry study involving 150,000 PBMCs from multiple donors before and after stimulation with BCR/FcR-XL \citep{bodenmiller2012cytof}.
-We used iSEE to vizualise and refine a gating analysis to obtain B cells, and to investigate differences in expression of the functional marker pS6 after stimulation.\\
+Finally, we explored a mass cytometry study involving more than 170,000 PBMCs from multiple donors before and after stimulation with BCR/FcR-XL \citep{bodenmiller2012cytof}.
+We used iSEE to visualise and refine a gating analysis to obtain B cells, and to investigate differences in expression of the functional marker pS6 after stimulation.\\
 
@@
 Any study that can be represented in a SummarizedExperiment object can be used as input, allowing iSEE to accommodate a diverse range of `omics data sets.
-The interface is flexible and can be dynamically customized by the user; supports exploration of interactions between data aspects through colouring and linking between panels;
+The interface is flexible and can be dynamically customised by the user; supports exploration of interactions between data aspects through colouring and linking between panels;
 and provides transparency and reproducibility during the interactive analysis, through code tracking and state reporting. 
@@
 \section*{Acknowledgements}
-We thank the organizers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
+We thank the organisers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
 We also thank members of the Bioconductor community for their helpful suggestions.

### 2018-05-10 manuscript/main.tex
Minor edits from John's comments.

@@
 iSEE is directly compatible with many existing R/Bioconductor packages for analysing high-throughput biological data,
-and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
+and provides useful features such as simultaneous examination of (meta)data and analysis results, dynamic linking between plots and code tracking for reproducibility.
 We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics data sets.
@@
 Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
-To this end, visualisation of the data in an intuitive and interactive interface is crucial for allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).\\
+To this end, visualisation of the data using an intuitive and interactive interface is crucial for allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).\\
 
@@
 For instance, individual assays may represent gene expression matrices, either in the form of raw counts or as values obtained by different normalisation strategies.
-In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'' slots, respectively; those may include experimental metadata as well as any analysis results.\\
+In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'' slots, respectively; these may include experimental metadata as well as any analysis results.\\
 
@@
 \subsection*{Using a multi-panel layout}
-All data aspects stored in a SummarizedExperiment can be readily examined alongside one another in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
+All data aspects stored in a SummarizedExperiment can be simultaneously examined in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
 The interface layout is built using the shinydashboard package \citep{chang2018shinydashboard}, using colour-coded panels to visualise each data aspect.
@@
 We also thank members of the Bioconductor community for their helpful suggestions.
+Finally, we thank John Marioni for helpful comments on the manuscript.
 

### 2018-05-14 manuscript/main.tex
Updated Makefile, added Mark to acknowledgements.
Added F1000Res editorial team changes.

@@
 \usepackage{listings}
+\usepackage{hyperref}
 
@@
 
-Most existing tools for interactive visualisation of biological data are designed for specific assays and analyses, e.g., pRoloc for proteomics \citep{gatto2014mass}, shinyMethyl for methylation \citep{fortin2014shinymethyl}, HTSvis for high-throughput screens \citep{scheeder2017htsvis}.
+Most existing tools for interactive visualisation of biological data are designed for specific assays and analyses, e.g., \href{https://bioconductor.org/packages/release/bioc/html/pRoloc.html}{pRoloc} for proteomics \citep{gatto2014mass}, \href{http://bioconductor.org/packages/release/bioc/html/shinyMethyl.html}{shinyMethyl} for methylation \citep{fortin2014shinymethyl}, \href{http://htsvis.dkfz.de/HTSvis/}{HTSvis} for high-throughput screens \citep{scheeder2017htsvis}.
 Opportunities for customisation are generally limited, making it difficult to re-use the same visualisation software for new technologies or experimental designs where different aspects of the data are of interest.
-Moreover, standalone tools such as the Loupe Cell Browser from 10x Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}.
+Moreover, standalone tools such as the \href{https://support.10xgenomics.com/single-cell-gene-expression/software/visualization/latest/what-is-loupe-cell-browser}{Loupe Cell Browser} from 10x Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}.
 This complicates any coordinated use of these tools with a reproducible, transparent, and statistically rigorous analysis.\\
@@
 Here, we present the iSEE software package for interactive data exploration.
-iSEE is implemented in R using the Shiny framework \citep{chang2017shiny} and exploits data structures from the open-source Bioconductor project \citep{gentleman2004bioconductor}, specifically the SummarizedExperiment class.
+iSEE is implemented in R using the \href{https://shiny.rstudio.com/}{Shiny framework}  \citep{chang2017shiny} and exploits data structures from the open-source \href{https://www.bioconductor.org/}{Bioconductor project}  \citep{gentleman2004bioconductor}, specifically the SummarizedExperiment class.
 iSEE allows users to simultaneously visualise multiple aspects of a given data set, including experimental data, metadata and analysis results.
@@
 We demonstrate the capabilities of iSEE by applying it to a diverse range of real data sets.
+
+\section*{Operation}
+% * <research@f1000.com> 2018-05-14T09:51:45.749Z:
+% 
+% > Operation
+% please complete this section regarding the minimal system requirements for your software tool and an overview of the work flow
+% 
+% ^.
 
@@
 All data aspects stored in a SummarizedExperiment can be simultaneously examined in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
-The interface layout is built using the shinydashboard package \citep{chang2018shinydashboard}, using colour-coded panels to visualise each data aspect.
+The interface layout is built using the \href{https://rstudio.github.io/shinydashboard/}{shinydashboard} package \citep{chang2018shinydashboard}, using colour-coded panels to visualise each data aspect.
 Individual panel types include:
@@
 \item Column data plots, for visualising sample metadata stored in the colData slot of the SummarizedExperiment object.
-\item Feature assay plots, for visualising expression values for a particular feature (e.g., gene) across samples from any assay in the SummarizedExperiment object.
+\item Feature assay plots, for visualising expression values for a particular feature (e.g. gene) across samples from any assay in the SummarizedExperiment object.
 \item Row statistics tables, to present the contents of the rowData slot for the SummarizedExperiment object.
@@
 \begin{figure*}[t]
-\includegraphics[width=\textwidth]{pics/Fig1.pdf}
+\includegraphics[width=\textwidth]{Fig1.pdf}
 \caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualises a different aspect of the data.
@@
 For these panel types, a scatter plot is automatically produced if the selected variables on the x- and y-axes are both continuous.
-If exactly one variable is categorical, points are grouped by the categorical levels and a (vertical or horisontal) violin plot is produced with points scattered within each violin.
+If exactly one variable is categorical, points are grouped by the categorical levels and a (vertical or horizontal) violin plot is produced with points scattered within each violin.
 If both variables are categorical, a ``rectangle plot'' is produced where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of that combination.
@@
 
-By default, colour maps for categorical and continuous variables are taken from the ggplot2 \citep{wickham2009ggplot2} and viridis packages \citep{garnier2018viridis}, respectively.
+By default, colour maps for categorical and continuous variables are taken from the \href{http://ggplot2.tidyverse.org/}{ggplot2}  \citep{wickham2009ggplot2} and \href{https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html}{viridis} packages \citep{garnier2018viridis}, respectively.
 However, iSEE also implements the ExperimentColorMap class, which allows users to specify arbitrary colour maps for particular variables.
@@
 \subsection*{Additional functionalities}
-Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl \citep{zerbino2018ensembl} or Entrez \citep{ncbi2017entrez}.
+Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as \href{https://www.ensembl.org/index.html}{Ensembl} \citep{zerbino2018ensembl} or  \href{https://www.ncbi.nlm.nih.gov/Class/MLACourse/Original8Hour/Entrez/}{Entrez} \citep{ncbi2017entrez}.
 % do we need to cite them?
@@
 For large data sets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualisation.
-Users can also include a bespoke step-by-step ``tour'' of their data set via the rintrojs package \cite{ganz2016rintrojs}, guiding the audience through an examination of the salient features in the data.
+Users can also include a bespoke step-by-step ``tour'' of their data set via the \href{https://introjs.com/}{rintrojs} package \cite{ganz2016rintrojs}, guiding the audience through an examination of the salient features in the data.
 
@@
 
-Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE/}
+An interactive tour of this use case can be viewed \href{http://shiny.imbei.uni-mainz.de:3838/iSEE/}{here}
 
@@
 
-Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_PBMC4k/}
+An interactive tour of this use case can be viewed \href{http://shiny.imbei.uni-mainz.de:3838/iSEE_PBMC4k/}{here}
 
@@
 
-Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_TCGA/}
+An interactive tour of this use case can be viewed \href{http://shiny.imbei.uni-mainz.de:3838/iSEE_TCGA/}{here}
 
@@
 
-Link: \url{http://shiny.imbei.uni-mainz.de:3838/iSEE_cytof/}
+An interactive tour of this use case can be viewed \href{http://shiny.imbei.uni-mainz.de:3838/iSEE_cytof/}{here}
 
@@
 
-\section*{Data and software availability}
+\section*{Software availability}
+% * <research@f1000.com> 2018-05-14T10:15:33.183Z:
+% 
+% > Data and software availability
+% please provide an archived version of you code and the licence under which it can be used, see email
+% 
+% ^.
 The iSEE package is available at \url{https://bioconductor.org/packages/iSEE}.
@@
 Code for the demonstrations and tours is available at \url{https://github.com/LTLA/iSEE2018}.
+% * <research@f1000.com> 2018-05-14T10:10:51.545Z:
+% 
+% The author contribution section has been deleted, since CRediT was filled in on submission and will be used for the Author Contributions section.
+% 
+% ^.
+\section*{Data availability }
 
-\section*{Author contributions}
-All authors contributed conceptually and technically to the development and maintenance of large sections of the package code.
-ATLL and FM developed the application and data infrastructure.
-KRA and CS developed the plotting framework.
-FM developed the code tracking and user interface.
-KRA designed and developed the ExperimentColorMap class.
-All authors contributed to testing the application on all data sets.
-All authors wrote and approved the final manuscript.
-
+Data used in the described use cases is available from the following articles: \\
+\url{http://doi.org/10.1038/nn.4216}\citep{tasic2016allen} \\
+\url{http://doi.org/10.1038/ncomms14049}\citep{zheng2017massively} \\
+\url{https://doi.org/10.1093/bioinformatics/btv377}\citep{piccolo2015TCGA} \\
+\url{https://doi.org/10.1038/nbt.2317}\citep{bodenmiller2012cytof} \\
 \section*{Competing interests}
@@
 Finally, we thank John Marioni for helpful comments on the manuscript.
+% * <research@f1000.com> 2018-05-14T10:11:46.639Z:
+% 
+% > John Marioni 
+% please confirm you have permission to name them in your manuscript 
+% 
+% ^.
 

### 2018-05-14 manuscript/main.tex
Made changes as requested by F1000Res.

@@
 \section*{Operation}
-% * <research@f1000.com> 2018-05-14T09:51:45.749Z:
-% 
-% > Operation
-% please complete this section regarding the minimal system requirements for your software tool and an overview of the work flow
-% 
-% ^.
+The iSEE software package requires R version 3.5.0 or higher, along with packages from Bioconductor version 3.7 or higher.
+The interface is initialised with a single call to the \texttt{iSEE()} function, accepting a SummarizedExperiment object \citep{huber2015orchestrating} as input.
+Any analysis workflow that generates a SummarizedExperiment object is supported.
 
-\section*{Taking a SummarizedExperiment input}
-The iSEE interface is initialised with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
+\section*{Motivation for using the SummarizedExperiment class}
 Each instance of the SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where rows and columns represent genomic features and biological samples, respectively.
@@
 \section*{Software availability}
-% * <research@f1000.com> 2018-05-14T10:15:33.183Z:
-% 
-% > Data and software availability
-% please provide an archived version of you code and the licence under which it can be used, see email
-% 
-% ^.
-The iSEE package is available at \url{https://bioconductor.org/packages/iSEE}.
+The iSEE package is available at \url{https://bioconductor.org/packages/iSEE} under the MIT license.
 Source code of the development version of the package is available at \url{https://github.com/csoneson/iSEE}.
 Code for the demonstrations and tours is available at \url{https://github.com/LTLA/iSEE2018}.
-% * <research@f1000.com> 2018-05-14T10:10:51.545Z:
-% 
-% The author contribution section has been deleted, since CRediT was filled in on submission and will be used for the Author Contributions section.
-% 
-% ^.
+
 \section*{Data availability }
-
 Data used in the described use cases is available from the following articles: \\
@@
 \url{https://doi.org/10.1038/nbt.2317}\citep{bodenmiller2012cytof} \\
+
 \section*{Competing interests}
@@
 We also thank members of the Bioconductor community for their helpful suggestions.
-Finally, we thank John Marioni for helpful comments on the manuscript.
-% * <research@f1000.com> 2018-05-14T10:11:46.639Z:
-% 
-% > John Marioni 
-% please confirm you have permission to name them in your manuscript 
-% 
-% ^.
+Finally, we thank John Marioni and Mark Robinson for their helpful comments on the manuscript.
 

### 2018-05-15 tours/tcga_app.R
downsampling resolution increased to 200

@@ -33,7 +33,7 @@ cd$ColorByColData <- "CancerType"
 cd$VisualChoices[[1]] <- c("Color", "Points", "Other")
 cd$PointAlpha <- 0.25
 cd$Downsample <- TRUE
-cd$SampleRes <- 100
+cd$SampleRes <- 200
 cd$LegendPosition <- "Right"
 # point selection
 cd$BrushData <- list(
@@ -68,7 +68,7 @@ rd$VisualChoices[[1]] <- c("Color", "Points", "Other")
 rd$ColorBy <- "Column data"
 rd$ColorByColData <- "CancerType"
 rd$Downsample <- TRUE
-rd$SampleRes <- 100
+rd$SampleRes <- 200
 rd$LegendPosition <- "Right"
 # select
 rd$SelectBoxOpen <- TRUE
@@ -86,7 +86,7 @@ fe$YAxisFeatName <- match("ERBB2", rownames(sce))
 fe$VisualBoxOpen <- TRUE
 fe$VisualChoices[[1]] <- c("Points")
 fe$Downsample <- TRUE
-fe$SampleRes <- 100
+fe$SampleRes <- 200
 # select
 fe$SelectBoxOpen <- TRUE
 fe$SelectByPlot <- c("Column data plot 1", "Column data plot 2")
